%\RequirePackage{pdfmanagement-testphase}
%\DeclareDocumentMetadata{pdfversion=1.7,lang=en-UK}

\documentclass{amsproc}

\usepackage{amsmath,amssymb,graphicx}
%\usepackage{amsmath,amssymb,accsupp,tagpdf,xstring,graphicx}
%\usepackage[tagpdf]{axessibility}


\newtheorem{theorem}{Theorem}[subsection]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{example}[theorem]{Example}
\newtheorem{xca}[theorem]{Exercise}

\newenvironment{solution}{\begin{proof}[Solution]}{\end{proof}}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\renewcommand{\Im}{\mathop{\textup{Im}}}
\renewcommand{\Re}{\mathop{\textup{Re}}}
\renewcommand{\d}{\partial}


\DeclareMathOperator{\N}{\mathbb{N}}
\DeclareMathOperator{\Z}{\mathbb{Z}}
\DeclareMathOperator{\R}{\mathbb{R}}
\DeclareMathOperator{\C}{\mathbb{C}}
\DeclareMathOperator{\cosec}{cosec}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\lcm}{lcm}
\DeclareMathOperator{\res}{\textup{res}}
\DeclareMathOperator{\sgn}{sgn}

\numberwithin{equation}{section}

\begin{document}

\title[Complex Analysis]{3U Complex Methods/3H Methods in Complex Analysis \\
Lecture Notes}
\author{Christian Voigt}

%Version 2022-23

\maketitle

% \begin{center}
% Christian Voigt
% \end{center}

\vspace{2cm}

This course is mainly concerned with the differentiation and integration of complex functions (complex-valued functions defined on subsets of the complex plane). Differentiation is in many ways the same as for real functions, but there are some interesting differences. In particular, a complex function differentiable on an open disc or on the entire plane~$\C$ can always be differentiated infinitely often and be expressed as the sum of a Taylor series. Integration has a much richer theory than for real functions, because there are many paths in a plane from one point to another.

These lecture notes are based on material prepared by Richard Steiner and Christian Korff.

\vfill
\pagebreak

%Week 1

\section*{PART I. DIFFERENTIATION}

\subsection{Limits}

Sequences of complex numbers and their limits are defined in basically the same way as for real numbers. By definition, a sequence of complex numbers is a
function $ \N \rightarrow \mathbb{C} $, usually written $ (z_n)_{n = 1}^\infty $ where $ z_n \in \mathbb{C} $.

We say that $ (z_n)_{n = 1}^\infty $ converges to $ z \in \C $ if for every $ \epsilon > 0 $ there exists $ n_0 \in \N $, such that
for any $ n \in \N $ with $ n \geq n_0 $, we have $ |z_n - z| < \epsilon $.
Equivalently,
$$
\forall\epsilon>0\;\exists n_0 \in \N \text{ such that } n > n_0 \Rightarrow |z_n - z| < \epsilon.
$$
In this case we shall write $ \lim_{n\to\infty}z_n = z $.

\begin{lemma}
A sequence $ (z_n)_{n = 1}^\infty $ of complex numbers converges to $ z \in \C $ if and only if the real sequences $ (\Re(z_n))_{n = 1}^\infty $
and $ (\Im(z_n))_{n = 1}^\infty $ are convergent with limits $ \Re(z) $ and $ \Im(z) $, respectively.
\end{lemma}

A particulary important role in complex analysis are played by infinite series. By definition, a series of complex numbers is an expression of the form
$$
S = \sum_{n = 0}^\infty a_n
$$
for complex numbers $ a_0, a_1, \dots $.
The series $ S $ is called convergent if the sequence $ (S_k)_{k = 0}^\infty $ of partial sums $ S_k = \sum_{n = 0}^k a_n $ converges.
In this case the limit $ \lim_{n\to\infty} S_n $ is called the sum of the series.
Many properties of real series, including criteria for checking convergence, carry over to the complex case.

\begin{example}
Let $ z \in \mathbb{C} $ be such that $ |z| < 1 $. Then the geometric series
$$
\sum_{n = 0}^\infty z^n
$$
converges, and its sum is $ \frac{1}{1 - z} $.
\end{example}

We shall also be interested in limits of functions.

\begin{definition}
Let $ U $ be a subset of~$ \C $, let $ f $ be a function from~$ U $ to~$ \C $, and let $ a,w \in \C $. We say that $ f(z) $ converges to $ w $ as $ z $ tends to $ a $ if

(1) there exists $ R > 0 $ such that
$$
0 < |z - a| < R \Rightarrow z \in U;
$$

(2) for all $ \epsilon > 0 $ there exists $ \delta > 0 $ such that if $ z \in U $ then
$$
0 < |z - a| < \delta \Rightarrow \left|f(z) - w\right| < \epsilon.
$$

In this case we write $ \lim_{z \to a} f(z) = w $.
\end{definition}

Note that $ a $ need not be contained in $ U $. Limits of functions can be described in terms of sequences, in analogy to the real case.

\begin{proposition}
Let $ U $ be a subset of~$ \C $, let $ f $ be a function from~$ U $ to~$ \C $, and let $ a \in \C $. Assume that there exists $ R > 0 $ such that
$ 0 < |z - a| < R \Rightarrow z \in U $.

Then $ f(z) $ converges to $ w $ as $ z $ tends to $ a $ if
and only if for every sequence $ (z_n)_{n = 1}^\infty $ in $ U $ with $ \lim_{n \to \infty} z_n = a $ and $ z_n \neq a $ for all $ n $
we have $ \lim_{n \to \infty} f(z_n) = w $.
\end{proposition}

\vfill
\pagebreak


\subsection{Introduction to differentiation} \label{CMS1}

Differentiation of complex functions is defined in essentially the same way as over the real numbers.

\begin{definition} \label{CMD1.1}
Let $ U $ be a subset of~$ \C $, let $ f: U \rightarrow \C $ be a function, and let $ a $ be a point of~$ U $.
One says that $ f $~is \emph{differentiable} at~$ a $, with \emph{derivative}~$ l $, if
\begin{enumerate}
\item[(1)] there exists $ r > 0 $ such that
$$
|z - a| < r \Rightarrow z \in U;
$$
\item[(2)] for all $ \epsilon > 0 $ there exists $ \delta > 0 $ such that if $ z \in U $ then
$$
0 < |z - a| < \delta \Rightarrow \left|\dfrac{f(z) - f(a)}{z - a} - l\right| < \epsilon.
$$
\end{enumerate}
\end{definition}

In other words, $ f $~is differentiable at~$ a $, with derivative~$ l $, if $ f $~is defined at all points sufficiently close to~$ a $ and if the difference quotient
$$
\dfrac{f(z) - f(a)}{z - a}
$$
converges to~$ l $ when $ z $~tends to~$ a $. As usual, if the derivative of~$ f $ at~$ a $ exists, then it is denoted $ f'(a) $. If $ f: U \rightarrow \C $
is differentiable at all points of $ U $ we also write
$$
\dfrac{d}{dz}f(z) \quad (\text{or } \dfrac{d}{dz}f = \dfrac{df}{dz})
$$
for the derivative, viewed as a function on $ U $ with values in $ \C $.

Many results regarding differentiability for complex functions are the same as for real functions.
For instance, if $ c $~is a constant then it is easy to see that
$$
\dfrac{dc}{dz} = 0.
$$
The following rules are verified in the same way as in real analysis.

\begin{lemma}
If $ f: U \rightarrow \C $~and~$ g: U \rightarrow \C $ are differentiable functions then
\begin{itemize}
\item (Additivity) $ \dfrac{d}{dz}(f\pm g)=\dfrac{df}{dz}\pm\dfrac{dg}{dz}, $
\item (Product Rule) $ \dfrac{d}{dz}(fg)=g\dfrac{df}{dz}+f\dfrac{dg}{dz}, $
\item (Quotient Rule) $ \dfrac{d}{dz}\dfrac{f}{g}=\dfrac{g\dfrac{df}{dz}-f\dfrac{dg}{dz}}{g^2} $ provided $ g $ is nowhere zero.
\end{itemize}
\end{lemma}

By using these results, one can differentiate polynomials and rational functions, just as one would expect.

\begin{example}
If $ n $~is a natural number, then
$$
\dfrac{d}{dz}z^n = nz^{n-1}.
$$
\end{example}

\begin{solution}
This follows inductively by writing $ z^n = z^{n - 1} z $ and using the product rule.
\end{solution}

The chain rule holds as well. That is, if $g$~is differentiable at~$a$ and $f$~is differentiable at $ g(a) $, then $ f\circ g $ is differentiable at~$ a $ and
$$
(f \circ g)'(a) = f'\bigl(g(a)\bigr)g'(a).
$$

%An equation
%$$F(z,w)=0$$
%can be differentiated implicitly near a solution $(z,w)=(a,b)$ treating~$w$ as a function of~$z$, provided that derivatives involved are continuous and that one does not divide by zero. More precisely, the equation determines~$w$ close to~$b$ uniquely for $z$ close to~$a$, and $w$~is differentiable with derivative given by
%$$\dfrac{\d F}{\d z}+\dfrac{\d F}{\d w}\dfrac{dw}{dz}=0,$$
%provided that $\d F/\d z$ and $\d F/\d w$ are continuous and that $\d F/\d w\neq 0$ at $(a,b)$.

%\begin{example} \label{CME1.2}
%Consider the equation
%$$w^2=z.$$
%Implicit differentiation gives
%$$2w\dfrac{dw}{dz}=1,$$
%hence
%$$\dfrac{dw}{dz}=\dfrac{1}{2w},$$
%provided that $w\neq 0$. Hence, if $a$~and~$b$ are such that $b^2=a$ and $b\neq 0$, then every number~$z$ close to~$a$ has a unique square root~$w$ close to~$b$, and $w$~is differentiable with respect to~$z$, and the derivative is as described.
%\end{example}

%This result could of course be obtained by using polar forms. The approach using implicit differentiation is easier to use and more general. It also evades the problems caused by the fact that a non-zero complex number has two square roots.

\begin{example} \label{CME1.3}
Show that the function~$ f(z) = \bar{z} $ is not differentiable anywhere on $ \C $.
\end{example}

\begin{solution}
Let $ a \in \C $ be arbitrary. Let $ z \in \C $ be a nearby point with $ |z - a| = r > 0 $. Then $ z = a + re^{i\theta} $ for some~$\theta$, and the difference quotient is
$$
\dfrac{\bar{z} - \bar{a}}{z - a} = \dfrac{\bar{a} + re^{-i\theta}-\bar{a}}{a + re^{i\theta} - a} = \dfrac{e^{-i\theta}}{e^{i\theta}} = e^{-2i\theta}.
$$
The difference quotient takes arbitrary values with modulus~$ 1 $ for arbitrarily small values of $ |z - a| $, so it does not tend to a limit as $ z $~tends to~$ a $.
Therefore the function $ z \mapsto \bar{z} $~is not differentiable at~$ a $.
\end{solution}

\begin{example} \label{CME1.4}
At which points is the function~$ |z|^2 $ differentiable?
\end{example}

\begin{solution}
Let $ a \in \C $ be any point and let $ z = a + re^{i\theta} $ be a nearby point. Then the difference quotient is
\begin{align*}
\dfrac{z\bar z - a\bar a}{z-a}
&= \dfrac{(a + re^{i\theta})(\bar{a} + re^{-i\theta}) - a\bar{a}}{re^{i\theta}} \\
&= \dfrac{are^{-i\theta} + \bar{a} re^{i\theta} + r^2}{re^{i\theta}} \\
&= ae^{-2i\theta} + \bar{a} + re^{-i\theta}.
\end{align*}
In this expression $\bar{a} + re^{-i\theta} $ tends to~$ \bar{a} $ as $ r $~tends to~$ 0 $. If $ a \neq 0 $, then $ ae^{-2i\theta} $ takes arbitrary values of modulus~$ |a| $
with $ z $~arbitrarily close to~$ a $, so the difference quotient does not tend to a limit. But if $ a = 0 $, then the difference quotient tends to~$ 0 $. Therefore $ |z|^2 $~is
differentiable only at $ z = 0 $.
\end{solution}

This is a curiosity; we are really interested in functions which are differentiable on large regions, not just at isolated points.

\begin{definition} \label{CMD1.5}
Let $ c $ be a point in~$ \C $ and let $ r $ be a positive real number. Then the \emph{open disc} with centre~$ c $ and radius~$ r $ is the set
$$
\{z \in \C \mid |z - c| < r \}.
$$
\end{definition}

\begin{definition} \label{CMD1.6}
A subset~$ U $ of~$ \C $ is \emph{open} if for every point~$ a \in U $  there is an open disc with centre~$ a $ entirely contained in~$ U $.
\end{definition}

\begin{example} \label{CME1.7}
Every open disc is open, and $ \C $~is open.
\end{example}

\begin{definition} \label{CMD1.8}
A \emph{holomorphic function} is a function $ f \colon U \to \C $ such that $ U $~is an open subset of~$ \C $ and $ f $~is differentiable at every point of~$ U $.
\end{definition}

This course is about holomorphic functions. Note that, for a complex function defined on an open set~$ U $, the first condition for differentiability in
Definition~\ref{CMD1.1} is automatically satisfied at every point of~$ U $.


\vfill\pagebreak

\subsection{Power series} \label{CMS2}

Complex series and power series behave just like real ones. By a complex power series we mean a series of the form
$$
\sum_{n = 0}^\infty a_n(z-c)^n = a_0 + a_1(z-c) + a_2(z-c)^2 + \cdots,
$$
where $ c $~and $ a_0, a_1, \dots $ are complex numbers. Let us list some important facts about such power series.

\begin{theorem} \label{power}
Let $ \sum_{n = 0}^\infty a_n(z - c)^n $ be a complex power series.
\begin{enumerate}
\item  There is a radius of convergence~$ R $ with $ 0 \leq R \leq \infty $ such that the series converges
for $ |z - c| < R $ and diverges for $ |z - c| > R $.
\item Suppose that $ R > 0 $ and let $ D $ be the open disc with centre~$ c $ and radius~$ R $, with the convention that $ D = \C $ if $ R = \infty $.
Let $ f(z) $ be the sum of the series. Then $ f $~is holomorphic on~$ D $. The derivative is given by
$$
f'(z) = \sum_{n = 1}^\infty na_n(z-c)^{n-1} = a_1 + 2a_2(z-c) + 3a_3(z-c)^2 + \cdots,
$$
and also has radius of convergence~$ R $.
\end{enumerate}
\end{theorem}

In many cases the radius of convergence can be found from the ratio test.

\begin{theorem}[Ratio Test] \label{CMT2.2}
If a power series
$$
\sum_{n=0}^\infty a_n(z - c)^n = a_0 + a_1(z - c) + a_2(z - c)^2 + \cdots
$$
is such that $ |a_{n + 1}/a_n| $ converges to a finite or infinite limit~ $ 0 \leq L \leq \infty $ as $ n \to \infty $, then the radius of convergence is $ 1/L $.
\end{theorem}

Here we use the convention $ 1/0 = \infty $ and $ 1/\infty = 0 $.

Assume now that $ f(z) = \sum_{n = 0}^\infty a_n(z - c)^n $ is a complex power series with radius of convergence $ R $, and let $ D $ be the open disc of radius $ R $ centred at $ c $.
By applying Theorem \ref{power} repeatedly, one sees that the function $ f $~can be differentiated infinitely often on $ D $ and that
$$
f^{(n)}(c) = n! a_n.
$$
The original series is then equal to the Taylor series
$$
f(z) = \sum_{n = 0}^\infty \dfrac{f^{(n)}(c)}{n!}(z - c)^n = f(c) + f'(c)(z - c) + \dfrac{f''(c)}{2!}(z - c)^2 + \dfrac{f'''(c)}{3!}(z - c)^3 + \cdots.
$$
There are also functions~$ F $ on~$ D $ such that $ F'(z) = f(z) $, in other words, we can always find an \emph{antiderivative} of~$ f $. These functions are given by the series
$$
F(z) = k + \sum_{n=0}^\infty\dfrac{a_n}{n+1}(z-c)^{n+1},
$$
where $ k $~is an arbitrary constant, and one can show that these series also have radius of convergence~$ R $.

\begin{example} \label{CME2.1}
From the theory of geometric series, there is a power series $ S(z) $ in~$ z $ with radius of convergence~$ 1 $ given by
$$
S(z) = \sum_{n=0}^\infty z^n = 1 + z + z^2 + \cdots
$$
such that $ S(z) = 1/(1 - z) $. By differentiating and integrating term by term, we obtain power series $ D(z) $ and $ I(z) $ with radius of convergence~$ 1 $ such that
\begin{align*}
D(z) &= 1 + 2z + 3z^2 + \cdots, \\
I(z) &= z + \tfrac{1}{2}z^2 + \tfrac{1}{3}z^3 + \cdots
\end{align*}
and such that
\begin{align*}
D(z) &= \dfrac{d}{dz} \dfrac{1}{1 - z} = - \dfrac{1}{(1 - z)^2} \dfrac{d}{dz}(1 - z) = \dfrac{1}{(1 - z)^2}, \\
I'(z) &= \dfrac{1}{1 - z}.
\end{align*}
By modifying this example, if $ c $~is any non-zero complex number, then $ 1/z $ has an antiderivative on the open disc with centre~$ c $ and radius~$ |c| $.
\end{example}


\vfill\pagebreak


%Week 2

\subsection{Elementary functions} \label{CMS3}

The \emph{exponential function} $ \exp(z) = e^z $ can be defined for all complex numbers~$ z $ as the sum of the power series
$$
\exp(z) = e^z = 1 + z + \dfrac{z^2}{2!} + \dfrac{z^3}{3!} + \cdots = \sum_{n = 0}^\infty \dfrac{z^n}{n!},
$$
the radius of convergence being infinite by the ratio test. One can then define $ \cos(z) $ and $ \sin(z) $ for all complex numbers~$ z $ by
$$
\cos(z) = \dfrac{e^{iz} + e^{-iz}}{2} = 1 - \dfrac{z^2}{2!} + \dfrac{z^4}{4!} - \cdots = \sum_{n = 0}^\infty (-1)^n \frac{z^{2n}}{(2n)!}
$$
and
$$
\sin(z) = \dfrac{e^{iz} - e^{-iz}}{2i} = z - \dfrac{z^3}{3!} + \dfrac{z^5}{5!} - \cdots = \sum_{n = 0}^\infty (-1)^n \frac{z^{2n + 1}}{(2n + 1)!} ,
$$
the radii of convergence again being infinite. These definitions agree with the usual definitions when $ z $~is real, and as in the real case one has
$$
\dfrac{d}{dz}e^z=e^z,\quad
\dfrac{d}{dz}\cos(z)=-\sin(z),\quad
\dfrac{d}{dz}\sin(z)=\cos(z).
$$
We note also that \emph{Euler's identity}
$$
e^{iz}
= \dfrac{e^{iz}+e^{-iz}}{2}+\dfrac{e^{iz}-e^{-iz}}{2}
= \cos(z) + i\sin(z)
$$
holds for all complex numbers~$ z $.

\begin{lemma} \label{CME3a}
For any fixed $ w \in \C $ the function $ z \mapsto e^{z + w}e^{-z} $ is constant.
\end{lemma}

\begin{proof}
For~$ w \in \C $ one has
$$
\dfrac{d}{dz}(e^{z+w} e^{-z}) = e^{z+w}e^{-z} + e^{z+w}(-e^{-z}) = 0,
$$
using the product rule and the above formula for the derivative of $ \exp $.
It follows that
$$
e^{z+w}e^{-z} = e^{0+w}e^{-0} = e^w
$$
for all $ z,w \in \C $, which yields the claim.
\end{proof}

In particular, Lemma \ref{CME3a} shows $ e^z e^{-z} = e^0 = 1 $, so $ e^z $~is non-zero for all $ z \in \C $, and
$$
e^{-z} = \dfrac{1}{e^z}.
$$
For all complex numbers $ z $~and~$ w $ it now follows from the formula $ e^{z + w} e^{-z} = e^w $ that
\begin{align*}
&e^{z+w} = e^z e^w,\\
&e^{z-w} = \dfrac{e^z}{e^w}.
\end{align*}
In particular for $ z = x + iy $ with $ x,y \in \R $ we have
$$
e^z = e^{x + iy} = e^x e^{iy} = e^x(\cos(y) + i\sin(y)),
$$
so that the complex exponential function can be expressed in terms of the real exponential and trigonometric functions. The complex sine and cosine functions can
then also be expressed in terms of the real exponential and trigonometric functions, as can the other trigonometric functions like
$$
\tan(z) = \dfrac{\sin(z)}{\cos(z)}.
$$
Moreover we can easily recover various known properties of trigonometric functions.

\begin{example} \label{CME3.0}
Show that $ \cos^2(z) + \sin^2(z) = 1 $ for all $ z \in \C $.
\end{example}

\begin{solution}
We calculate
\begin{align*}
\cos^2(z) + \sin^2(z) &= (\cos(z) + i\sin(z))(\cos(z) - i\sin(z)) \\
&= \bigg(\frac{e^{iz} + e^{-iz}}{2} + \frac{e^{iz} - e^{-iz}}{2}\bigg) \bigg(\frac{e^{iz} + e^{-iz}}{2} - \frac{e^{iz} - e^{-iz}}{2}\bigg) \\
&= e^{iz} e^{-iz} = 1,
\end{align*}
directly from the definition of $ \sin $ and $ \cos $.
\end{solution}

We also have the hyperbolic sine and cosine functions defined by
$$
\cosh(z) = \tfrac{1}{2}(e^z + e^{-z}), \quad \sinh(z) = \tfrac{1}{2}(e^z - e^{-z}),
$$
and other hyperbolic functions like $ \tanh $. These hyperbolic functions are just slight variants of the usual trigonometric functions,
in the sense that
$$
\cosh(z) = \cos(iz), \quad \sinh(z) = -i\sin(iz), \quad \tanh(z) = -i \tan(iz)
$$
for all $ z \in \C $.

\begin{example} \label{CME3.0.5}
Show that $ \cosh^2(z) - \sinh^2(z) = 1 $ for all $ z \in \C $.
\end{example}

\begin{solution}
This is similar to Example \ref{CME3.0}. We calculate
$$
\cosh^2(z) - \sinh^2(z) = (\cosh(z) + \sinh(z))(\cosh(z) - \sinh(z)) = e^z e^{-z} = 1,
$$
directly from the definition of the hyperbolic sine and cosine functions.
\end{solution}

From Example \ref{CME3.0.5} it follows in particular that the points $ (\cosh(t), \sinh(t)) \in \R^2 $ for $ t \in \R $ form the unit hyperbola
$$
H = \{(x,y) \in \R^2 \mid x^2 - y^2 = 1 \text{ and } x > 0 \}.
$$
This is the reason why these functions are called hyperbolic. By comparison, note that the set of all $ (\cos(t), \sin(t)) $ for $ t \in \R $ is the unit circle, which is
why the functions $ \sin $ and $ \cos $ are called trigonometric.

Let us next discuss \emph{logarithms}. For a given complex number~$ z $ we consider the equation
$$
e^w = z.
$$
The solutions~$ w \in \C $ of this equation are called the complex logarithms of~$ z $. If $ z = 0 $ then there are no solutions, since we have seen above that $ e^w $
is nonzero for every $ w \in \C $. For $ z \neq 0 $ let
$$
z = r e^{i\theta} \quad (r,\theta \in \R,\ r>0)
$$
be a polar form of~$ z $, and let
$$
w = u + iv\quad (u,v \in \R)
$$
be the cartesian form of~$ w $; then
\begin{align*}
e^w = z
&\iff e^u e^{iv} = re^{i\theta}\\
&\iff u = \log(r),\ v = \theta + 2k\pi\quad (k\in\Z).
\end{align*}
We write this in the form
$$
\log(z) = \log|z| + i\arg(z).
$$
Here
%$ \log(z) $ means a value of the complex logarithm of~$ z $, where
$ \log |z| $ is the ordinary real natural logarithm of the positive real number~$ |z| $,
and $ \arg(z) $ is the argument of~$ z $.

Note that $ \arg(z) $ has infinitely many values: if $ \theta $~is one value then the complete set of values
is given by $ \theta + 2k\pi $ where $ k $~is any integer. Correspondingly, $ \log(z) $ has infinitely many values. The notation $ \log(z) $ should therefore be used with great care.

%It is convenient to write $ \Arg(z) $ for a specific branch of the argument, for instance, the unique value of the argument
%between $ -\pi $ and $ \pi $, or between $ 0 $ and $ 2\pi $. If we fix the principal branch $ -\pi < \arg(z) < \pi $ then we can view $ \log(z) = \log|z| + i \arg(z) $ as
%a function on the open set $ \C \setminus \{x \in \R \mid x \leq 0 \} $. This is called the principal branch of the logarithm.

\begin{example} \label{CME3.1}
Find the complex logarithms of~$ z $ for $ z = -1 + i\sqrt{3} $.
\end{example}

\begin{solution}
We have
$$
\log(z) = \log |z| + i\arg(z).
$$
For $ z = x + iy $ with $ x,y \in \R $ we have $ |z| = \sqrt{x^2 + y^2} $ and
$$
\arg(z) =
\begin{cases}
\tan^{-1}(y/x)+2k\pi&(x>0),\\
\tan^{-1}(y/x)+(2k+1)\pi&(x<0),
\end{cases}
$$
where $ k $~is an arbitrary integer. In this case $ |z| = 2 $ and
$$
\arg(z) = -\tfrac{1}{3}\pi + (2k+1)\pi,
$$
so
$$
\log(z) = \log(2) - \tfrac{1}{3}\pi i + (2k + 1)\pi i,
$$
where $ k $~is an arbitrary integer.
\end{solution}

If we restrict attention to some fixed branch of the argument, like $ -\pi < \theta < \pi $ we can turn $ \log $ into a (well-defined) holomorphic function.
The equation
$$
z = e^{\log(z)}
$$
can then be differentiated
%near any particular solution
to give
$$
1 = e^{\log z} \dfrac{d}{dz}\log(z) = z \dfrac{d}{dz} \log(z),
$$
hence
$$
\dfrac{d}{dz}\log(z) = \dfrac{1}{z}
$$
for all $ z \neq 0 $. In this way we get antiderivatives for the function $ 1/z $, at least locally.

We use logarithms to define generalised powers~$ z^w $ for $ z,w \in \C $ with $ z \neq 0 $ by the formula
$$
z^w = e^{\log(z) w}.
$$
These are again multivalued, so should be treated very cautiously. However, they do give holomorphic functions on small regions if one makes a continuous choice for the logarithm.
In general, one gets an infinite sequence of possible values for $ z^w $, but if $ w $~is a rational number $ m/n $ such that $ m $~and~$ n $ are coprime integers and $ n > 0 $,
then one gets precisely $ n $~distinct values. In particular, if $ m $~is an integer then $ z^m $~has a single value. These single values are as one would expect:
$$
\dots,\ z^{-2} = 1/(zz),\ z^{-1} = 1/z,\  z^0 = 1,\ z^1 = z,\ z^2 = zz, \dots.
$$

\begin{example} \label{CME3.2}
Calculate all values for $ (2i)^{1/2} $.
\end{example}

\begin{solution}
We have $ (2i)^{1/2} = e^{\tfrac{1}{2}\log(2i)} $, and $ \log(2i) = \log(2) + \frac{\pi}{2}i + 2k\pi i $ for $ k $ any integer. Therefore we obtain
$$
(2i)^{1/2} = e^{\log(2)/2 + \pi/4 i + k \pi i} = \pm \sqrt{2} e^{\pi/4 i} = \pm (1 + i)
$$
as solutions.
\end{solution}

Using the above methods we can also solve equations involving functions which are built out of the exponential function, like trigonometric functions.

\begin{example} \label{CME3.2}
Find the complex numbers~$ z $ such that $ \cos(z) = 10 $.
\end{example}

\begin{solution}
We treat the equation
$$
\tfrac{1}{2}(e^{iz} + e^{-iz}) = 10
$$
as a quadratic equation in~$ e^{iz} $. After rearrangement it becomes
$$
(e^{iz})^2-20 e^{iz} + 1 = 0.
$$
The solutions are
$$
e^{iz} = \dfrac{20\pm\sqrt{396}}{2} = 10\pm\sqrt{99},
$$
so that
$$
iz = \log(10\pm\sqrt{99}) + 2k\pi i,
$$
that is,
$$
z = -i\log(10\pm\sqrt{99}) + 2k\pi,
$$
where $ k $~is an arbitrary integer.
\end{solution}

\vfill\pagebreak


%Week 3

\subsection{The Cauchy--Riemann equations} \label{CMS4}

By taking real and imaginary parts, a complex-valued function of a complex variable can be regarded as a complex-valued function of two real variables or as a pair
of real-valued functions of two real variables. It turns out that the differentiability of the complex-valued function is essentially equivalent to two simultaneous
equations between the partial derivatives of the real-valued functions.

\begin{theorem} \label{CMT4.1}
Let $ w $ be a complex-valued function of a complex variable~$ z $ and let
$$
u = \Re(w),\ v = \Im(w),\ x = \Re(z),\ y = \Im(z),
$$
so that
$$
w(z) = w(x + iy) = u(x,y) + i v(x,y).
$$
If $ w $~is differentiable with respect to~$ z $, then the real-valued functions $ u $~and~$ v $ have (continuous) partial derivatives with respect to $ x $~and~$ y $ and
\begin{align*}
&\dfrac{dw}{dz} = \dfrac{\d w}{\d x} = \dfrac{\d u}{\d x} + i\dfrac{\d v}{\d x}, \\
&\dfrac{\d u}{\d x} = \dfrac{\d v}{\d y}, \\
&\dfrac{\d u}{\d y} = -\dfrac{\d v}{\d x}.
\end{align*}
Conversely, if $ u $~and~$ v $ are real-valued functions having continuous partial derivatives with respect to $ x $~and~$ y $ such that
the \emph{Cauchy--Riemann equations}
$$
\dfrac{\d u}{\d x} = \dfrac{\d v}{\d y}, \quad \dfrac{\d u}{\d y} = -\dfrac{\d v}{\d x}
$$
hold, then $ w $~is differentiable with respect to~$ z $.
\end{theorem}

\begin{proof}
Suppose first that $ w $~is differentiable with respect to~$ z $. Then for any given value of~$ z $ we have
\begin{align*}
\dfrac{dw}{dz} &= \lim_{h \to 0} \dfrac{w(z + h) - w(z)}{h} \\
&= \lim_{h \to 0} \dfrac{u(z + h) - u(z)}{h} + i\lim_{h \to 0} \dfrac{v(z + h) - v(z)}{h}.
\end{align*}
Restricting~$ h $ to real values shows that
$$
\dfrac{dw}{dz} = \dfrac{\d w}{\d x} = \dfrac{\d u}{\d x} + i\dfrac{\d v}{\d x}.
$$
Restricting~$ h $ to purely imaginary values shows that
$$
\dfrac{dw}{dz}
= \dfrac{1}{i}\dfrac{\d w}{\d y}
= \dfrac{1}{i}\dfrac{\d u}{\d y} + \dfrac{\d v}{\d y}
= -i\dfrac{\d u}{\d y} + \dfrac{\d v}{\d y}.
$$
Equating real and imaginary parts in the two expressions for $ dw/dz $ therefore gives
$$
\dfrac{\d u}{\d x} = \dfrac{\d v}{\d y}, \quad \dfrac{\d u}{\d y}=-\dfrac{\d v}{\d x}
$$
as required.

Conversely, suppose that $ u $~and~$ v $ have continuous partial derivatives satisfying the Cauchy--Riemann equations. There are then continuous real-valued
functions $ s $~and~$ t $ of~$ z $ such that
$$
s = \dfrac{\d u}{\d x} = \dfrac{\d v}{\d y},\quad t = \dfrac{\d u}{\d y} = -\dfrac{\d v}{\d x}.
$$
For any given value of~$ z $ we will show that
$$
\dfrac{w(z + h) - w(z) - h[s(z) - it(z)]}{h} \to 0
$$
as $ h \to 0 $; this will show that $ w $~is differentiable as a function of~$ z $, with derivative $ s - it $.

Let $ h = k + il $ with $ k,l \in \R $. Take $ w(z + h) - w(z) $ in the form
$$
[w(z + k + il) - w(z + il)] + [w(z + il) - w(z)]
$$
and expand this to obtain
$$
[u(z + k + il) - u(z + il)] + i[v(z + k+ il) - v(z + il)] + [u(z + il) - u(z)] + i[v(z +il) - v(z)].
$$
By the mean value theorem from real analysis we find $ z_1,z_2,z_3,z_4 $ such that
$$
w(z + h) - w(z) = ks(z_1) - ikt(z_2) + lt(z_3) + ils(z_4)
$$
where $ z_1 $~and~$ z_2 $ are between $ z + k + il $ and $ z + il $, and $ z_3 $~and~$ z_4 $ are between $ z + il $ and~$ z $.
Note here that if $ k = 0 $ we may take $ z_1 = z_2 = z + il $, and similarly $ z_3 = z_4 = z $ if $ l = 0 $.
The points $ z_1,z_2,z_3,z_4 $ depend on~$ h $, and they tend to~$ z $ as $ h \to 0 $. On the other hand we have
$$
h[s(z) - it(z)] = (k + il)[s(z) - it(z)] = ks(z) - ikt(z) + lt(z) + ils(z).
$$
Therefore
\begin{align*}
&\dfrac{w(z + h) - w(z) - h[s(z) - it(z)]}{h} \\
&= \dfrac{k}{h}[s(z_1) - s(z)]
-\dfrac{ik}{h}[t(z_2) - t(z)]
+\dfrac{l}{h}[t(z_3) - t(z)]
+\dfrac{il}{h}[s(z_4) - s(z)].
\end{align*}
Here $ |k/h| \leq 1 $ because $ k = \Re(h) $. Moreover we have $ s(z_1) - s(z) \to 0 $ as $ h \to 0 $ because $ s $~is continuous, and similarly for the other
terms in this expression. It follows that
$$
\dfrac{w(z + h) - w(z) - h[s(z) - it(z)]}{h} \to 0
$$
as $ h \to 0 $, as required.

This completes the proof.
\end{proof}

\begin{example} \label{CME4.2}
Let
$$
u = e^x\cos(y),\ v = e^x\sin(y).
$$
Then $ u $~and~$ v $ have continuous partial derivatives with respect to $ x $~and~$ y $, and
\begin{align*}
&\dfrac{\d u}{\d x} = e^x\cos(y) = \dfrac{\d v}{\d y}, \\
&\dfrac{\d u}{\d y} = -e^x\sin(y) = -\dfrac{\d v}{\d x}, \\
\end{align*}
so there is a holomorphic function given by
$$
f(x + iy) = e^x(\cos(y) + i\sin(y)) \quad (x,y\in\R)
$$
with
$$
f'(x + iy) = \dfrac{\d}{\d x} f(x + iy) = f(x + iy).
$$
This gives an alternative approach to the complex exponential function.
\end{example}

\begin{example} \label{CME4.3}
Let $ u = x $ and $ v = -y $; then $ \d u/\d x = 1 \neq -1 = \d v/\d y $, so the function
$$
x + iy \mapsto x - iy \quad (x,y\in\R)
$$
is not differentiable. This gives an alternative proof that the complex conjugate function is not holomorphic.
\end{example}

\vfill\pagebreak


\subsection{Harmonic functions} \label{CMS5}

A \emph{harmonic function} is a real-valued function~$ u $ of two real variables $ x $~and~$ y $ which satisfies the \emph{Laplace equation}
$$
\dfrac{\d^2 u}{\d x^2} + \dfrac{\d^2 u}{\d y^2} = 0.
$$
Harmonic functions are important in certain branches of applied mathematics.

\begin{theorem} \label{CMT5.1}
If $ u $~and~$ v $ satisfy the Cauchy--Riemann equations and have continuous second-order partial derivatives, then they are harmonic.
\end{theorem}

\begin{proof}
Using the Cauchy--Riemann equations and the symmetry of continuous second-order partial derivatives, we see that
\begin{align*}
\dfrac{\d^2 u}{\d x^2} + \dfrac{\d^2 u}{\d y^2} &= \dfrac{\d}{\d x}\dfrac{\d u}{\d x}+\dfrac{\d}{\d y}\dfrac{\d u}{\d y} \\
&= \dfrac{\d}{\d x}\dfrac{\d v}{\d y} - \dfrac{\d}{\d y}\dfrac{\d v}{\d x} \\
&= \dfrac{\d^2 v}{\d x\d y} - \dfrac{\d^2 v}{\d y\d x} \\
&= 0,
\end{align*}
and similarly for $ v $.
\end{proof}

Let $ u,v $ be a pair of real-valued functions with continuous second-order partial derivatives. If $ u $~and~$ v $ satisfy the Cauchy--Riemann equations,
then $ v $~is called a \emph{harmonic conjugate} of~$ u $. On reasonable domains, like open discs,
%on any simply connected open set, see https://en.wikipedia.org/wiki/Harmonic_conjugate
%The logarithm on $ \C \setminus \{0\} $ shows that this is necessary, we have $ log(z) = \log|z| + i \arg(z) $ locally, so that the real part is $ \log|z| $.
every harmonic function has a harmonic conjugate, so that harmonic functions are essentially the same as the real parts of holomorphic functions.

One can find a harmonic conjugate by solving the two Cauchy--Riemann equations successively.

\begin{example} \label{CME5.2}
Show that the function $ u = x^3 - 3x y^2 $ is harmonic, and find a harmonic conjugate.
\end{example}

\begin{solution}
We have
$$
\dfrac{\d u}{\d x} = 3x^2 - 3y^2, \quad \dfrac{\d u}{\d y} = -6xy.
$$
It follows that $ u $~is harmonic, because
$$
\dfrac{\d^2 u}{\d x^2} + \dfrac{\d^2 u}{\d y^2} = 6x - 6x = 0.
$$
For $ v $ to be a harmonic conjugate, we require
$$
\dfrac{\d v}{\d x} = -\dfrac{\d u}{\d y}, \quad \dfrac{\d v}{\d y} = \dfrac{\d u}{\d x},
$$
that is,
$$
\dfrac{\d v}{\d x} = 6xy, \quad \dfrac{\d v}{\d y} = 3x^2 - 3y^2.
$$
The general solution to the first of these equations is
$$
v = 3x^2 y + \phi(y),
$$
where $ \phi(y) $ is an arbitrary function in~$ y $. For $ v $~to be a solution to the second equation one then requires $ \phi $ to be differentiable and
$$
3x^2 + \phi'(y) = 3x^2 - 3y^2,
$$
which means that $ \phi(y) = -y^3 + k $, where $ k $~is constant. Thus the harmonic conjugates~$ v $ are given by
$$
v = 3x^2y - y^3 + k,
$$
where $ k $~is constant.

Note that when performing the first antidifferentiation in the above argument one needs $ \phi $ to be an arbitrary function, not just an arbitrary constant.
\end{solution}

This example shows that there is a holomorphic function~$ f $ given by
$$
f(z) = f(x + iy) = w = u + iv = (x^3 - 3xy^2) + i(3x^2y - y^3 + k).
$$
In fact, by the binomial theorem this can be written as $ f(z) = z^3 + ik $.

Heuristically, one can often find a formula for $ f(z) $ from the formula for $ f(x + iy) $ by replacing~$ x $ with~$ z $
and replacing~$ y $ with~$ 0 $; this is because
$$
\dfrac{dw}{dz} = \dfrac{\d w}{\d x}.
$$

\vfill\pagebreak


\section*{PART II. INTEGRATION}

\subsection{Introduction to integration} \label{CMS6}

An important part of real analysis concerns the integration of functions over closed intervals of the real line. Over the complex numbers we can integrate
functions in a similar way, but we now have additional flexibility in choosing the domains we want to integrate over.

We will be interested in the contour integration of functions in the complex plane. Essentially, if $ w = u + iv $ is a function of $ z = x + iy $ with $ u,v,x,y $
real and if $ \gamma $~is a path in the complex plane then we will get
$$
\int_\gamma w\,dz = \int_\gamma (u + iv)(dx +i\,dy) = \int_\gamma (u\,dx - v\,dy) + i\int_\gamma (v\,dx + u\,dy),
$$
which means that we are dealing with pairs of real line integrals. We will, however, give the definitions independently in a reasonably general form.

\begin{definition} \label{CMD6.1}
Let $ U $ be a subset of~$ \C $. A \emph{path} in~$ U $ is a continuous map $ \gamma: [a,b] \to U, $ where $ [a,b] $ is a closed interval in~$ \R $.
\end{definition}

Let $ \gamma: [a,b] \to U $ be a path. We can approximate the length of~$ \gamma $ by sums of the form
$$
\sum_{r = 1}^n |\gamma(a_r) - \gamma(a_{r - 1})|,
$$
where
$$
a = a_0 < a_1 < \cdots < a_n = b.
$$
If these sums tend to a finite limit~$ L $ as $ \max_r(a_r - a_{r - 1}) $ tends to zero, then one says that $ \gamma $~has \emph{finite length}~$ L $.

If $ \gamma $~is of finite length and if $ f: U \to \C $ is a continuous function, then the \emph{contour integral}
$$
\int_\gamma f(z)\,dz
$$
is obtained from the \emph{Riemann sums}
$$
\sum_{r = 1}^n f\bigl(\gamma(t_r)\bigr)[\gamma(a_r) - \gamma(a_{r-1})],
$$
where
$$
a = a_0 < a_1 < \cdots < a_n = b
$$
is a partition of $ [a,b] $, and where
$$
a_{r - 1} \leq t_r \leq a_r,
$$
by taking the limit as $ \max_r(a_r - a_{r - 1}) \to 0 $.

We will not give the details of this construction in this course. In the sequel, the paths involved in contour integrals are supposed to be of finite length, even if this is not stated.

Note that the family of Riemann sums is determined by the \emph{contour} of $ \gamma $, that is the points in the image of~$ \gamma $, and by the order in which they occur;
the precise parametrisation does not matter.
Because of this, we tend to specify paths by stating their images and the direction of travel. For example, we may refer to an integral along a straight line segment
from~$ P $ to~$ Q $ or an anticlockwise circular arc; this really means the segment or arc with any convenient parametrisation in the correct direction. Note that the direction
of travel must be specified, because of the following result.

\begin{lemma} \label{CMP6.2}
If the direction of travel along a path is reversed, then the integrals along the path are multiplied by~$ -1 $.
\end{lemma}

\begin{proof}
The Riemann sums for the reverse direction are obtained from the original Riemann sums by multiplying by~$ -1 $.
\end{proof}

There is a very important and very useful bound for contour integrals, as follows.

\begin{proposition} \label{CMT6.3}
If $ \gamma $~is of finite length~$ L $ and if $ |f(z)| \leq M $ for all points ~$ z $ in the image of~$ \gamma $, then
$$
\left|\int_\gamma f(z)\,dz\right| \leq ML.
$$
\end{proposition}

\begin{proof}
For each Riemann sum one has
\begin{align*}
\left|\sum_{r = 1}^n f\bigl(\gamma(t_r)\bigr)[\gamma(a_r) - \gamma(a_{r - 1})]\right|
&\leq \sum_{r = 1}^n |f\bigl(\gamma(t_r)\bigr)|\,|\gamma(a_r) - \gamma(a_{r - 1})| \\
&\leq M\sum_{r = 1}^n |\gamma(a_r) - \gamma(a_{r - 1})|,
\end{align*}
and the result follows by taking the limits as $ \max_r(a_r - a_{r - 1}) $ tends to zero.
\end{proof}

This result can be used to estimate integrals. The ability to do this is an intended learning objective for Methods in complex analysis but not for Mathematics 3U, and this
is the only real difference in the intended learning objectives between the two courses. But the result should be known in any case, because it is frequently used in proofs.

In practice, contour integrals are converted into integrals over intervals in the real line, essentially by change of variables.

\begin{definition} \label{CMD6.4}
A path~$ \gamma $ on $ [a,b] $ is a \emph{piecewise $ C^1 $~path} if $ \gamma $ continuously differentiable at all but finitely many points of the interval $ [a,b] $.
\end{definition}

\begin{proposition} \label{CMP6.5}
Let $ \gamma: [a,b] \to U $ be a piecewise $ C^1 $~path. Then $ \gamma $~has finite length if and only if the absolute value of derivative $ |\gamma'| $~has a finite integral
over $ [a,b] $. If $ \gamma $~has finite length, then the length is given by
$$
\int_a^b |\gamma'(t)|\,dt.
$$
If $ \gamma $~has finite length and if $ f $~is a continuous function on~$ U $, then
$$
\int_\gamma f(z)\,dz = \int_a^b f\bigl(\gamma(t)\bigr)\,\gamma'(t)\,dt.
$$
\end{proposition}
%If $ \gamma:[-1,1] \to \C, \gamma(t) = t + i t^2 \sin(1/t^2) $ with $ \gamma(0) = (0,0) $ then $ \gamma $ is piecewise C^1 (on (-1,0) and (0,1)) but not rectifiable:
%The derivative becomes $ \gamma'(t) = 1 + i 2t \sin(1/t^2) - 2 t^{-1} \cos(1/t^2) $ and $ \int_0^1 t^{-1} \cos(1/t^2) dt = -\int_1^\infty \tau^{-1} \cos(\tau) d\tau $ with $ \tau = 1/t $
%and $ dt = - 1/\tau^2 d\tau $ does not converge.

We will not give a proof of this result. Let us instead take a look at some examples.

\begin{example} \label{CME6.6}
Let $ \gamma: [a,b] \to \C $ be the path given by
$$
\gamma(\theta) = c + re^{i\theta}
$$
for some $ c \in \C $ and $ r > 0 $. Compute the length of~$ \gamma $.
\end{example}

\begin{solution}
We have $ \gamma'(\theta) = ire^{i\theta} $, so the length of~$ \gamma $ is
$$
\int_a^b |i re^{i\theta}|\,d\theta = \int_a^b r\,d\theta = r(b - a).
$$
\end{solution}

\begin{example} \label{CME6.7}
Let $ \gamma $ be the straight line from $ 1 + i $ to $ 2 + 3i $. Compute
$$
\int_\gamma \bar{z}\,dz.
$$
\end{example}

\begin{solution}
We can parametrise~$ \gamma $ as
$$
\gamma(t) = (1 + i) + [(2 + 3i) - (1 + i)]t = (1 + i) + (1 + 2i)t
$$
for $ 0 \leq t \leq 1 $. Then $ \overline{\gamma(t)} = (1 - i) + (1 - 2i)t $ and $ \gamma'(t) = 1 + 2i $, so
$$
\int_\gamma \bar{z}\,dz = \int_0^1 [(1 - i) + (1 - 2i)t](1 + 2i)\,dt
= \int_0^1 (3 + i + 5t)\,dt
= [(3 + i)t + \tfrac{5}{2}t^2]_0^1
= \tfrac{11}{2} + i.
$$
Informally, we may write $ z = (1 + i) + (1 + 2i)t $ and substitute $ dz = (1 + 2i)\,dt $ and $ \bar{z} = (1 - i) + (1 - 2i)t $ in $ \int_\gamma \bar{z} dz $
to calculate the integral.
\end{solution}

In general the straight line from~$ p $ to~$ q $ can be parametrised as
$$
z = p + t(q - p)
$$
for $ 0 \leq t \leq 1 $. Other parametrisations are possible; in particular one does not have to use the interval $ [0,1] $.

\begin{example} \label{CME6.8}
Let $ \gamma: [a, b] \to \C $ be the path given by
$$
\gamma(t) = e^{it},
$$
and let $ n $ be an integer. Compute
$$
\int_\gamma z^n\,dz.
$$
\end{example}

\begin{solution}
We have
\begin{align*}
\int_\gamma z^n\,dz &= \int_a^b \gamma(t)^n\,\gamma'(t)\,dt \\
&= \int_a^b e^{int}ie^{it}\,dt \\
&= \int_a^b ie^{i(n + 1)t}\,dt.
\end{align*}
In cases with $ n \neq -1 $ we get
$$
\int_\gamma z^n\,dz = \left[\dfrac{e^{i(n + 1)t}}{n + 1}\right]_a^b = \dfrac{e^{i(n + 1)b} - e^{i(n + 1)a}}{n + 1},
$$
and we get
$$
\int_\gamma z^{-1}\,dz = \int_a^b i\,dt = i(b - a)
$$
in the case $ n = -1 $.
\end{solution}

In order to calculate contour integrals we can often use what amounts to the fundamental theorem of calculus.

\begin{theorem}[Fundamental Theorem of Calculus] \label{CMP6.9}
Suppose $ \gamma $~is a
%piecewise $ C^1 $
path in an open set $ U \subset \C $ from~$ c $ to~$ d $ and let $ f $ be a continuous function on~$ U $ such that $ f(z) = F'(z) $
for some differentiable function~$ F $ on~$ U $, then
$$
\int_\gamma f(z)\,dz = [F(z)]_c^d = F(d) - F(c).
$$
\end{theorem}

This gives another way to look at Example~\ref{CME6.8} in the case $ n \neq -1 $: since $ \gamma $~is a path from~$ e^{ia} $ to~$ e^{ib} $ in $ U = \mathbb{C} \setminus \{0\} $
we have
$$
\int_\gamma z^n\,dz = \left[\dfrac{z^{n + 1}}{n + 1}\right]_{e^{ia}}^{e^{ib}} = \dfrac{e^{i(n + 1)b} - e^{i(n + 1)a}}{n + 1}.
$$

\begin{definition} \label{CMD6.10}
A \emph{closed path} in $ U \subset \C $ is a path~$ \gamma: [a,b] \to \C $ such that $ \gamma(a) = \gamma(b) $.
\end{definition}

\begin{proposition} \label{CMP6.11}
If $ \gamma $~is a closed
%piecewise $ C^1 $
path in~$ U $ and if $ f $~is a continuous function on~$ U $ such that $ f(z) = F'(z) $ for some differentiable function~$ F $ on~$ U $, then
$$
\int_\gamma f(z)\,dz = 0.
$$
\end{proposition}

\begin{example} \label{CME6.12}
If $ \gamma $~is a closed
%piecewise $ C^1 $
path in~$ \C $ and if $ f $~is a polynomial, then
$$
\int_\gamma f(z)\,dz=0.
$$
\end{example}

\begin{example} \label{CME6.13}
Let $ \gamma $ be the circle with centre~$ 0 $ and radius~$ 1 $ traced anticlockwise. By integrating along~$ \gamma $, show that there is no function on $ \C \setminus \{0\} $
with derivative~$ z^{-1} $.
\end{example}

\begin{solution}
We can parametrise~$ \gamma $ as $ z = e^{it} $ for $ 0 \leq t \leq 2\pi $. We have
$$
dz = ie^{it}\,dt = iz\,dt,
$$
hence
$$
\int_\gamma z^{-1}\,dz = \int_0^{2\pi} z^{-1}iz\,dt = \int_0^{2\pi} i\,dt = 2\pi i\neq 0.
$$
Since $ \gamma $~is a closed path, it follows that $ z^{-1} $~cannot have an anti-derivative on $ \C \setminus \{0\} $.
\end{solution}

\vfill\pagebreak


\subsection{Cauchy's theorem} \label{CMS7}

By the fundamental theorem of calculus, the integral along a closed path of a function with an anti-derivative is equal to zero, see Proposition~\ref{CMP6.11}.
Cauchy's theorem says that, under certain circumstances, one gets the same result for functions with derivatives rather than anti-derivatives.

We begin with the special case in which the domain of the closed curve is the boundary of a rectangle.

\begin{theorem}[Cauchy's theorem for images of rectangles] \label{CMT7.1}
Let $ f: U \to \C $ be a differentiable function on an open set~$ U $ in~$ \C $, let $ R = [a,b] \times [a',b'] $ be a rectangle in~$ \R^2 $ with boundary $ \d R $,
and let $ \gamma: \d R \to U $ be a path of finite length such that $ \gamma = \Gamma|\d R $ for some continuous function $ \Gamma: R \to U $. Then
$$
\int_\gamma f(z)\, dz = 0.
$$
\end{theorem}

\begin{proof}[Sketch of proof]
The function~$ \Gamma $ may be very far from smooth. However, since $ U $~is open and $ \gamma $~is of finite length, we have enough wriggle room to change~$ \Gamma $ without
changing its values on the boundary such that it multiplies lengths by a bounded amount. We assume that this has been done. In particular, let $ R' $~is a rectangle contained
in~$ R $ with boundary $ \d R' $ traced in the same direction as~$ \gamma $; then $ \Gamma|\d R' $ is of finite length and we can make the definition
$$
I(R') = \int_{\Gamma|\d R'} f(z)\,dz.
$$
In this notation, we want to show that $ I(R) = 0 $.

Bisect~$ R $ horizontally and vertically into four rectangles $ S,T,U,V $; then
$$
I(R) = I(S) + I(T) + I(U) + I(V)
$$
because the integrals along edges inside~$ R $ cancel out. We can therefore choose one of the small rectangles, $ R_1 $~say, such that
$$
|I(R)| \leq 4 \,|I(R_1)|.
$$
By repeating this process, we can construct a sequence of rectangles
$$
R \supset R_1 \supset R_2 \supset \cdots,
$$
each a quarter of its predecessor, such that
$$
|I(R)| \leq 4^n\,|I(R_n)|.
$$

Now let $ (s_n,t_n) $ be the lower left hand corner of~$ R_n $. The sequence $ s_1,s_2\,\dots\, $ is increasing and bounded above, so it converges to a limit~$ s $, and
similarly $ t_n $~converges to a limit~$ t $. Let $ w = \Gamma(s,t) $; then $ w \in \Gamma(R_n) $ for all~$ n $.

We now use the fact that $ f $~is differentiable. For $ z \in U \setminus\{w\} $ let
$$
q(z) = \dfrac{f(z) - f(w)}{z - w} - f'(w),
$$
and let $ q(w) = 0 $; then $ q $~is continuous on~$ U $ and
$$
f(z) = f(w) + f'(w)(z - w) + q(z)(z - w)
$$
for all $ z \in U $. Since the integral of a polynomial along a closed curve is zero (see Example~\ref{CME6.12}) we have
$$
I(R_n) = \int_{\Gamma|\d R_n} q(z)(z - w)\,dz.
$$

Now the linear dimensions of~$ R_n $ are obtained from those of~$R$ by multiplying by~$ 2^{-n} $. Since $ \Gamma $~multiplies lengths by a bounded amount, there are positive
constants $ K $~and~$ L $ such that the length of $ \Gamma|\d R_n $ is at most $ 2^{-n}K $ and such that $ |z - w| \leq 2^{-n}L $ for $ z $ in $ \Gamma|\d R_n $.
We also have $ q(z) \to 0 $ as $ z \to w $, so there are positive constants $ M_n $ such that $ |q(z)| \leq M_n $ for $ z $ in $ \Gamma|\d R_n $ and such
that $ M_n \to 0 $ as $ n \to \infty $. By Proposition~\ref{CMT6.3},
$$
|I(R_n)| \leq 2^{-n}K2^{-n}LM_n = 4^{-n}KLM_n.
$$
Since $ |I(R)| \leq 4^n\,|I(R_n)| $ and since $ M_n \to 0 $ as $ n \to \infty $, it follows that $ I(R) = 0 $. This completes the proof.
\end{proof}

We apply this result to show that in certain circumstances the integrals of a function along different paths are equal.

\begin{definition} \label{CMD7.2}
Let
$$
\gamma,\delta:[a,b] \to U
$$
be two closed paths with the same domain $ [a,b] $ and the same codomain~$ U $. Then $ \gamma $~and~$ \delta $ are \emph{homotopic} (as closed paths in~$ U $) if there is a
continuous function
$$
\Gamma: [0,1] \times [a,b] \to U
$$
such that
\begin{align*}
&\Gamma(0,t) = \gamma(t) \quad (a \leq t \leq b),\\
&\Gamma(1,t) = \delta(t) \quad (a \leq t \leq b),\\
&\Gamma(s,a) = \Gamma(s,b) \quad (0 \leq s \leq 1).
\end{align*}
The function~$ \Gamma $ is called a \emph{homotopy} from~$ \gamma $ to~$ \delta $.
\end{definition}

In this definition, the homotopy~$ \Gamma $~is a continuous family of closed paths $ \Gamma(s,-) $, one for each point~$ s $ in $ [0,1] $, starting with $ \gamma $ and
finishing with~$ \delta $.

\begin{theorem}[Homotopy invariance] \label{CMT7.3}
Let $ f: U \to \C $ be a differentiable function on an open set~$ U $ in~$ \C $ and let $ \gamma $~and~$ \delta $ be closed paths in~$ U $ of finite length which are homotopic as
closed paths in~$ U $. Then
$$
\int_\gamma f(z)\,dz = \int_\delta f(z)\,dz.
$$
\end{theorem}

\begin{proof}[Sketch of proof]
By a wriggle room argument, we can find a homotopy whose restriction to the boundary of its domain is of finite length. The integral of~$ f $ along its boundary in the appropriate direction is then
$$
\int_\gamma f(z)\,dz - \int_\delta f(z)\,dz
$$
since the integrals along the other two sides cancel out because the homotopy is a homotopy through closed paths. The result therefore follows from Theorem~\ref{CMT7.1}.
\end{proof}

We use this to generalise Theorem~\ref{CMT7.1} as follows. Let us say that a closed path in an open set $ U \subset \C $ is \emph{null-homotopic} in $ U $ if it is homotopic to a
constant path as a closed path in~$ U $. Then the following result is an easy consequence of Theorem~\ref{CMT7.3} because an integral along a constant path is obviously
equal to zero.

\begin{theorem}[Cauchy's theorem for null-homotopic paths] \label{CMT7.4}
Let $ f: U \to \C $ be a differentiable function on an open set~$ U $ in~$ \C $ and let $ \gamma $ be a closed path in~$ U $ which is null-homotopic in~$ U $. Then
$$
\int_\gamma f(z)\,dz = 0.
$$
\end{theorem}

In particular we can apply this result to paths which do not cross themselves.

\begin{definition} \label{CMD7.5}
A \emph{simple closed path} on an interval $ [a,b] $ is a closed path~$ \gamma $ such that $ \gamma(t_1) \neq \gamma(t_2) $ for $ a \leq t_1 < t_2 < b $.
\end{definition}

Thus a simple closed path is a closed path which is as near to being injective as possible. Note that for a closed path $ \gamma $ we have $ \gamma(a) = \gamma(b) $ by definition,
which means that a closed path cannot be injective.

Let $ \gamma $ be a simple closed path in~$ \C $. Then the complement of~the image of~$ \gamma $ is the union of
two disjoint non-empty connected open sets, one of them bounded, one of them unbounded; this is the famous \emph{Jordan curve theorem}. The members of the bounded set
are \emph{inside}~$ \gamma $, and the members of the unbounded set are \emph{outside}~$ \gamma $. If $ U $~is an open set containing~$ \gamma $ and all the points
inside~$ \gamma $, then one can show $ \gamma $~is homotopic to a constant path in~$ U $. As a special case of Theorem~\ref{CMT7.4} we therefore get the following result.

\begin{theorem}[Cauchy's theorem] \label{CMT7.6}
Let $ \gamma $ be a simple closed path of finite length in~$ \C $, let $ U $ be an open set containing~$ \gamma $ and all the points inside~$ \gamma $, and let $ f: U \to \C $ be differentiable. Then
$$
\int_\gamma f(z)\,dz = 0.
$$
\end{theorem}

%The main results to remember in this section are Theorems \ref{CMT7.3} and~\ref{CMT7.6}. From Theorem~\ref{CMT7.3} it is easy to deduce Theorem~\ref{CMT7.4} and then
%Theorem~\ref{CMT7.1}.

\vfill\pagebreak


\subsection{Cauchy's integral formulae} \label{CMS8}

The theorems in the previous section are very powerful. We will now describe some of their consequences.

If $ \gamma $ is a simple closed path in~$ \C $ then $ \gamma $~may be traced anticlockwise or clockwise. To be more precise, let $ U $ be a set containing~$ \gamma $ and all
the points inside~$ \gamma $, and let $ z $ be a point inside~$ \gamma $. One can show that $ \gamma $~is homotopic in $ U \setminus \{z\} $ to a small circle with centre~$ z $,
and one can show that the circles for the various points~$ z $ are all traced the same way, either anticlockwise or clockwise. We say that $ \gamma $~itself is traced anticlockwise
or clockwise accordingly. We will mainly consider paths that are traced anticlockwise, because of the following theorem.

\begin{theorem}[Cauchy's integral formula] \label{CMT8.1}
Let $ \gamma $ be a simple closed path in~$ \C $ traced anticlockwise, let $ U $ be an open set containing~$ \gamma $ and all the points inside~$ \gamma $, let $ f: U \to \C $ be differentiable, and let $ z $ be a point inside~$ \gamma $. Then
$$
f(z) = \dfrac{1}{2\pi i} \int_\gamma \dfrac{f(\zeta)}{\zeta - z}\,d\zeta.
$$
\end{theorem}

\begin{proof}[Sketch of proof]
If we write
$$
I = \int_\gamma \dfrac{f(\zeta)}{\zeta - z}\,d\zeta
$$
then we must show that $ I = 2\pi i f(z) $.

Let $ \delta $ be a circle with centre~$ z $ and with small radius~$ r $ traced anticlockwise. If $ r $~is sufficiently small, then $ \gamma $~is homotopic to~$ \delta $
in $ U \setminus \{z\} $. The function $ \zeta \mapsto f(\zeta)/(\zeta - z) $ is differentiable on $ U \setminus \{z\} $, so
$$
I = \int_\delta \dfrac{f(\zeta)}{\zeta - z}\,d\zeta
$$
by Theorem~\ref{CMT7.3}.

We now consider what happens if $ f(\zeta) $ is replaced by $ f(z) $ in the integral~$ I $. We can parametrise~$ \delta $ as $ \zeta = z + re^{it} $ with $ t $ running
from~$ 0 $ to $ 2\pi $, and we get
$$
\int_{\delta} \dfrac{f(z)}{\zeta - z}\,d\zeta = \int_0^{2\pi} \dfrac{f(z)}{re^{it}}ire^{it}\,dt = 2\pi if(z).
$$
It follows that
$$
I - 2\pi i f(z) = \int_\delta \dfrac{f(\zeta) - f(z)}{\zeta - z}\,d\zeta.
$$
Now the difference quotient $ [f(\zeta) - f(z)]/(\zeta - z) $ converges to $ f'(z) $ as $ \zeta $~tends to~$ z $; therefore, if $ r $~is sufficiently small we get
$$
\left|\dfrac{f(\zeta) - f(z)}{\zeta - z}\right| \leq |f'(z)| + 1
$$
for all~$ \zeta $ in~$ \delta $. The length of~$ \delta $ is $ 2\pi r $, so for such~$ r $ it follows from Proposition~\ref{CMT6.3} that
$$
|I - 2\pi i f(z)| \leq 2 \pi r\bigl(|f'(z)| + 1\bigr).
$$
Since $ r $~can be arbitrarily small we conclude that $ I = 2 \pi i f(z) $ as required.
\end{proof}

In this theorem, the integrand $ f(\zeta)/(\zeta - z) $ is infinitely often differentiable with respect to~$ z $ for each fixed~$ \zeta $ in the path~$ \gamma $.
One can deduce that $ f $~is infinitely often differentiable inside~$ \gamma $, with the obvious derivatives. This gives the following generalisation of Theorem \ref{CMT8.1}.

\begin{theorem}[Cauchy's integral formula for derivatives] \label{CMT8.2}
Let $ \gamma $ be a simple closed path in~$ \C $ traced anticlockwise, let $ U $ be an open set containing~$ \gamma $ and all the points inside~$ \gamma $, and
let $ f: U \to \C $ be differentiable. Then $ f $~can be differentiated infinitely often at any point~$ z $ inside~$ \gamma $, and the derivatives for $ n \geq 0 $ are given by
$$
f^{(n)}(z) = \dfrac{n!}{2\pi i}\int_\gamma \dfrac{f(\zeta)}{(\zeta - z)^{n + 1}}\,d\zeta.
$$
\end{theorem}

For every point~$ z $ in an open set~$ U $ there is a disc in~$ U $ with centre~$ z $. By considering a circle in this disc with centre~$ z $, one gets the following result.

\begin{theorem} \label{CMT8.3}
Let $ U $ be an open set in~$ \C $ and let $ f: U \to \C $ be a differentiable function. Then $ f $~can be differentiated infinitely often.
\end{theorem}

This is a very significant difference between complex analysis and real analysis. In fact, in real analysis there are differentiable functions whose derivative
is not even continuous, let alone differentiable. A basic example of this phenomenon is the function $ f: \R \to \R $ given by
$$
f(x) =
\begin{cases}
x^2 \sin(1/x) & x \neq 0 \\
0 & x = 0.
\end{cases}
$$
This function is differentiable in the sense of real analysis with
$$
f'(x) =
\begin{cases}
2x \sin(1/x) - \sin(1/x) & x \neq 0 \\
0 & x = 0.
\end{cases}
$$
This can be calculated using the product rule and chain rule as usual if $ x \neq 0 $, and directly from the definition at $ x = 0 $.
%https://math.stackexchange.com/questions/112067/how-discontinuous-can-a-derivative-be
Note that $ f' $ is discontinuous at $ 0 $.

\vfill\pagebreak


\subsection{Applications} \label{CME9}

We consider holomorphic functions on open discs and on the entire complex plane~$ \C $. These functions are not merely infinitely differentiable, as shown in Theorem~\ref{CMT8.3};
they are equal to the sums of their Taylor series.

In general, if $ U \subset \C $ is an open set then a function $ f: U \to \C $ is called analytic if it can be locally expressed as a sum of convergent power series.
We have already seen in Theorem \ref{power} that analytic functions are holomorphic. Now we shall establish the converse.

\begin{theorem} \label{CMT9.1}
Let $ f $ be a holomorphic function on an open set $ U \subset \C $, let $ c \in U $ and let $ D \subset U $ be an open disc with centre~$ c $.
Then $ f $ is analytic and
$$
f(z) = \sum_{n = 0}^\infty \dfrac{f^{(n)}(c)}{n!}(z - c)^n
$$
for all $ z $ in~$ D $.
\end{theorem}

\begin{proof}[Sketch of proof]
Let
$$
S = \sum_{n = 0}^\infty \dfrac{f^{(n)}(c)}{n!}(z - c)^n;
$$
we must show that $ S = f(z) $ for $ z \in D $. Let $ \gamma $ be a circle
%of radius $ r $
in~$ D $ with centre~$ c $ traced anticlockwise such that $ z $~is inside~$ \gamma $. From Cauchy's integral
formula for derivatives, see Theorem~\ref{CMT8.2}, one can deduce that
$$
S = \dfrac{1}{2\pi i} \sum_{n = 0}^\infty \int_\gamma \dfrac{f(\zeta)(z - c)^n}{(\zeta - c)^{n + 1}}\,d\zeta
= \dfrac{1}{2\pi i} \int_\gamma \sum_{n = 0}^\infty \dfrac{f(\zeta)(z - c)^n}{(\zeta - c)^{n + 1}}\,d\zeta,
$$
here the interchange of integration and summation is justified because one can check that the series
$$
\sum_{n = 0}^\infty \int_\gamma \left|\dfrac{f(\zeta)(z - c)^n}{(\zeta - c)^{n + 1}}\right| d\zeta
$$
%We want to apply Fubini: need to check that $ \sum \int |h(\zeta)| d\zeta $ converges. But the integrand is bounded by $ K q^n $ on $ \gamma $ where $ q = |z - c|/r < 1 $
%and independent of $ t $ where $ \gamma(t) = r e^{2 \pi i t} $, hence the integral just a factor 2 \pi i $
is convergent.
%, uniformly for $ |\zeta| \leq r $.

Consider now the expression
$$
\int_\gamma \sum_{n = 0}^\infty \dfrac{f(\zeta)(z - c)^n}{(\zeta - c)^{n + 1}}\,d\zeta = \int_\gamma \sum_{n = 0}^\infty \dfrac{f(\zeta)}{\zeta - c}
\left(\dfrac{z - c}{\zeta - c}\right)^n\,d\zeta.
$$
Since $ z $~is inside~$ \gamma $, we have
$$
\left|\dfrac{z - c}{\zeta - c}\right| < 1
$$
for each~$ \zeta $ on~$ \gamma $, so the integrand is an absolutely convergent geometric series with sum given by
$$
\sum_{n = 0}^\infty \dfrac{f(\zeta)(z - c)^n}{(\zeta - c)^{n + 1}}
= \dfrac{\left[\dfrac{f(\zeta)}{\zeta - c}\right]}{\left[1 - \dfrac{z - c}{\zeta - c}\right]}
= \dfrac{f(\zeta)}{\zeta - z}.
$$
Therefore
$$
S = \dfrac{1}{2\pi i} \int_\gamma \dfrac{f(\zeta)}{\zeta - z}\,d\zeta,
$$
and Cauchy's integral formula in Theorem~\ref{CMT8.1} implies that $ S = f(z) $ as required.
\end{proof}

Theorem \ref{power} and Theorem \ref{CMT9.1} show that a function $ f: U \to \C $ defined on an open set $ U \subset \C $ is analytic if and only if it is holomorphic.
This is why these two notions are often used interchangeably.

For holomorphic functions on the entire complex plane~$ \C $ there is a further key result due to Liouville, given as follows.

\begin{theorem}[Liouville's theorem] \label{CMT9.2}
Every non-constant holomorphic function on the entire complex plane~$ \C $ is unbounded.
\end{theorem}

\begin{proof}
Let $ f: \C \to \C $ be a bounded holomorphic function, say
$$
|f(z)| \leq M
$$
for all $ z \in \C $. We will show that $ f $~is constant by showing that $ f'(z) = 0 $ for all $ z \in \C $.

Indeed, let $ \gamma $ be a circle with centre~$ z $ and radius~$ r $ traced anticlockwise. By Cauchy's integral formula for derivatives, see Theorem~\ref{CMT8.2},
we have
$$
f'(z) = \dfrac{1}{2\pi i} \int_\gamma \dfrac{f(\zeta)}{(\zeta - z)^2}\,d\zeta.
$$
If $ \zeta $ is a point on~$ \gamma $ then
$$
\left|\dfrac{f(\zeta)}{(\zeta - z)^2}\right| = \dfrac{|f(\zeta)|}{r^2} \leq \dfrac{M}{r^2}.
$$
Since the length of~$ \gamma $ is $ 2\pi r $, it follows from Proposition~\ref{CMT6.3} that
$$
|f'(z)| \leq \left(\dfrac{1}{2 \pi}\right) \left(\dfrac{M}{r^2}\right)(2\pi r) = \dfrac{M}{r}.
$$
This is true for all positive numbers~$ r $, so $ f'(z) = 0 $ as required.
\end{proof}

%Students of Methods in complex analysis, but not students of Mathematics~3U, are intended to able to reproduce this proof and to construct analogous proofs for other results.

\begin{example} \label{CME9.3}
Let $ f $ be an holomorphic function on~$ \C $ such that $ \Re(f(z)) \geq 0 $ for all $ z \in \C $. Show that $ f $~is constant.
\end{example}

\begin{solution}
Let $ g(z) = e^{-f(z)} $; then $ g $~is a holomorphic function on~$ \C $ such that $ |g(z)| \leq 1 $ for all $ z \in \C $. By Liouville's theorem, $ g $~is constant.
Since $ f $~is continuous, $ f $~is constant.
\end{solution}

\vfill\pagebreak


\subsection{Laurent series} \label{CMS10}

Let $ c $ be a point in~$ \C $, let $ U $ be an open disc with centre~$ c $ or let $ U = \C $, let $ f $ be a holomorphic function on~$ U $, and let $ \gamma $ be a circle
in~$ U $ with centre~$ c $ traced anticlockwise. By Cauchy's integral formula for derivatives, Theorem~\ref{CMT8.2}, we have
$$
f^{(n)}(c) = \dfrac{n!}{2\pi i} \int_\gamma\dfrac{f(\zeta)}{(\zeta - c)^{n + 1}}\,d\zeta
$$
for all $ n \geq 0 $. The Taylor series
$$
f(z) = \sum_{n = 0}^\infty \dfrac{f^{(n)}(c)}{n!}(z - c)^n,
$$
which is valid for all~$ z $ in~$ U $ by Theorem~\ref{CMT9.1}, therefore takes the form
$$
f(z) = \sum_{n = 0}^\infty a_n(z - c)^n
$$
with
$$
a_n = \dfrac{1}{2\pi i} \int_\gamma\dfrac{f(\zeta)}{(\zeta - c)^{n + 1}}\,d\zeta.
$$
In this form, there is a simple generalisation to functions~$ f $ which are holomorphic only on $ U \setminus\{c\} $: we use the terms $ a_n(z - c)^n $ for all integers~$ n $,
both positive and negative. The resulting series is like a power series, except that it involves both positive and negative powers. Series of this kind are
called \emph{Laurent series;} thus a Laurent series is a series of the form
$$
\sum_{n = -\infty}^\infty a_n(z - c)^n.
$$

One can write a Laurent series as the sum of a power series in $ z - c $ and a power series in $ 1/(z - c) $,
$$
\sum_{n = -\infty}^\infty a_n(z - c)^n = \sum_{n = 0}^\infty a_n(z - c)^n + \sum_{m = 1}^\infty a_{-m}[1/(z - c)]^m;
$$
by definition, the Laurent series is convergent if and only if these two power series are both convergent.

\begin{theorem} \label{CMT10.1}
Let $ c $ be a point in~$ \C $, let $ U $ be an open disc with centre~$ c $ or let $ U = \C $, let $ f $ be a holomorphic function on $ U \setminus \{c\} $, let $ \gamma $ be
a circle in~$ U $ with centre~$ c $ traced anticlockwise, and let $ z $ be a point in $ U \setminus \{c\} $. Then
$$
f(z) = \sum_{n = -\infty}^\infty a_n(z - c)^n,
$$
where
$$
a_n = \dfrac{1}{2\pi i} \int_\gamma \dfrac{f(\zeta)}{(\zeta - c)^{n + 1}}\,d\zeta
$$
for all $ n \in \Z $.
\end{theorem}

\begin{proof}[Sketch of proof]
Let
\begin{align*}
S^+ &= \dfrac{1}{2\pi i} \int_\gamma \sum_{n = 0}^\infty \dfrac{f(\zeta)(z - c)^n}{(\zeta - c)^{n + 1}}\,d\zeta, \\
S^- &= \dfrac{1}{2\pi i} \int_\gamma \sum_{m = 0}^\infty \dfrac{f(\zeta)(z - c)^{-m - 1}}{(\zeta - c)^{-m}}\,d\zeta;
\end{align*}
we must show that
$$
S^+ + S^- = f(z).
$$

To evaluate~$ S^+ $, let $ \gamma^+ $ be a circle in~$ U $ with centre~$ c $ traced anticlockwise such that $ z $~is inside~$ \gamma^+ $. Then $ \gamma $~is homotopic
to~$ \gamma^+ $ in $ U \setminus \{c\} $; hence, by Theorem~\ref{CMT7.3},
$$
S^+ = \dfrac{1}{2\pi i} \int_{\gamma^+} \sum_{n = 0}^\infty \dfrac{f(\zeta)(z - c)^n}{(\zeta - c)^{n + 1}}\,d\zeta.
$$
As in the proof of Theorem~\ref{CMT9.1}, it follows that
$$
S^+ = \dfrac{1}{2\pi i} \int_{\gamma^+} \dfrac{f(\zeta)}{\zeta - z}\,d\zeta.
$$

To evaluate~$ S^- $, let $ \gamma^- $ be a circle in~$ U $ with centre~$ c $ traced anticlockwise such that $ z $~is \emph{outside}~$ \gamma^- $. We now get
$$
S^- = \dfrac{1}{2\pi i} \int_{\gamma^-} \sum_{m = 0}^\infty \dfrac{f(\zeta)(z - c)^{-m - 1}}{(\zeta - c)^{-m}}\,d\zeta
= \dfrac{1}{2\pi i} \int_{\gamma^-}\sum_{m = 0}^\infty \dfrac{f(\zeta)(\zeta - c)^m}{(z - c)^{m + 1}}\,d\zeta.
$$
The series in the integrand is a convergent geometric series because $ z $~is outside~$ \gamma^- $, so that
$$
\left|\dfrac{\zeta - c}{z - c}\right| < 1
$$
for all~$ \zeta $ on~$ \gamma^- $. The sum is now given by
$$
\sum_{m = 0}^\infty \dfrac{f(\zeta)(\zeta - c)^m}{(z - c)^{m + 1}} = \dfrac{\left[\dfrac{f(\zeta)}{z - c}\right]}{\left[1 - \dfrac{\zeta - c}{z - c}\right]}
= \dfrac{f(\zeta)}{z - \zeta}
= -\dfrac{f(\zeta)}{\zeta - z},
$$
so that
$$
S^-= -\dfrac{1}{2\pi i} \int_{\gamma^-} \dfrac{f(\zeta)}{\zeta - z}\,d\zeta.
$$

Recall now from Proposition~\ref{CMP6.2} that changing the sign of a contour integral is equivalent to reversing the direction of travel. Let $ \sigma $ be a closed path in~$ U $ consisting of the circle~$ \gamma^+ $ traced anticlockwise, a path~$ \lambda $ from a point on~$ \gamma^+ $ to a point on~$ \gamma^- $, the circle~$ \gamma^- $ traced clockwise, and
the path~$ \lambda $ traced in reverse; then we see that
$$
S^+ + S^- = \dfrac{1}{2\pi i} \int_\sigma \dfrac{f(\zeta)}{\zeta - z}\,d\zeta.
$$
We now shrink this path slightly to get an anticlockwise simple closed path~$\tau$ homotopic to~$ \sigma $ in $ U \setminus \{c,z\} $ such that $ z $~is inside~$ \tau $.
By Theorem~\ref{CMT7.3},
$$
S^+ + S^- = \dfrac{1}{2\pi i} \int_\tau \dfrac{f(\zeta)}{\zeta - z}\,d\zeta;
$$
hence, by Cauchy's integral formula, see Theorem~\ref{CMT8.1}, we get $ S^+ +S^- = f(z) $ as required.
\end{proof}

\vfill\pagebreak


\subsection{Singularities} \label{CMS11}

Let $ f $ be a holomorphic function on $ U \setminus \{c\} $, where $ U \subset \C $~is an open set containing ~$ c $. In this case one says that $ f $ has
an \emph{isolated singularity} at~$ c $. By Theorem~\ref{CMT10.1} there is a Laurent series expansion
$$
f(z) = \sum_{n = -\infty}^\infty a_n(z - c)^n
$$
for all~$ z $ sufficiently close to~$ c $.

\begin{definition} \label{CMD11.1}
With $ f $ and $ c $ as above there are the following three mutually exclusive possibilities.
\begin{enumerate}
\item Suppose that $ a_n = 0 $ for all $ n < 0 $. Then the series is a power series in the ordinary sense, with a non-zero radius of convergence. In particular it has the sum~$ a_0 $
at $ z = c $. One can therefore extend~$ f $ to a holomorphic function on~$ U $ by putting $ f(c) = a_0 $. One says that $ f $~has a \emph{removable singularity} at~$ c $.
\item Suppose that $ a_n $~is non-zero for for some negative values of~$ n $, but only for finitely many, so that the Laurent series has the form
$$
\sum_{n = -N}^\infty a_n(z - c)^n = (z - c)^{-N}[a_{-N} + a_{-N + 1}(z - c) + \cdots\,]
$$
with $ N > 0 $ and $ a_{-N} \neq 0 $. One says that $ f $~has a \emph{pole} of order~$ N $ at~$ c $. In this case,
$$
f(z) = \dfrac{g(z)}{(z - c)^N}
$$
for some function~$ g $ which is holomorphic on~$ U $ and non-zero at~$ c $.
\item Suppose that $ a_n $~is non-zero for infinitely many negative values of~$ n $. Then $ f $~has an \emph{isolated essential singularity} at~$ c $.
\end{enumerate}
\end{definition}

Let us consider examples for all three types of isolated singularities as introduced in Definition \ref{CMD11.1}.

\begin{example} \label{CME11.4}
The function
\begin{align*}
f(z) = \dfrac{\sin(z)}{z}
&= \dfrac{1}{z}\left(z - \dfrac{z^3}{3!} + \dfrac{z^5}{5!} - \cdots\,\right) \\
&= 1 - \dfrac{z^2}{3!} + \dfrac{z^4}{5!} - \cdots
\end{align*}
has a removable singularity at $ z = 0 $ and can be made holomorphic at~$ 0 $ by giving it the value~$ 1 $ at that point. We have $ f(z) \to 1 $ as $ z \to 0 $, so $ |f(z)| $
is bounded on some neighbourhood of~$ 0 $.

The function
$$
g(z) = \dfrac{z + 2}{(z - 3)(z + 5)^2}
$$
has a pole of order~$ 1 $ (a \emph{simple pole}) at $ z = 3 $, and a pole of order~$ 2 $ (a \emph{double pole}) at $ z = -5 $. Obviously $ |g(z)| \to \infty $
when $ z \to 3 $ or $ z \to -5 $.

The function
$$
h(z) = e^{1/z} = 1 + z^{-1} + \dfrac{z^{-2}}{2!} + \cdots
$$
has an isolated essential singularity at $ z = 0 $. If $ z \to 0 $ through positive real values, then $ |h(z)| \to \infty $; if $ z \to 0 $ through negative real values
then $ |h(z)| \to 0 $. For $ z $~tending to~$ 0 $ through arbitrary complex numbers, $ |h(z)| $ is not bounded but does not tend to infinity either.
\end{example}

Singularities can always be classified by the behaviour of the absolute value as in these examples.

\begin{theorem} \label{CMT11.5}
Let $ U \subset \C $ be an open set and let $ f $ be a holomorphic function on $ U \setminus \{c\} $, so that $ f $ has an isolated singularity at~$ c $.
\begin{enumerate}
\item The singularity is removable if and only if $ f $~is bounded on some neighbourhood of~$ c $.
\item The singularity is a pole if and only if $ |f(z)| \to \infty $ as $ z \to c $.
\item The singularity is an isolated essential singularity if and only if $ f $~is unbounded on any neighbourhood of~$ c $ and $ |f(z)| $ does not tend to infinity
as $ z $~tends to~$ c $.
\end{enumerate}
\end{theorem}

\begin{proof}
(1) If the singularity is removable, then obviously $f$~is bounded on some neighbourhood of~$c$.

Conversely, suppose that $ f $~is bounded on some neighbourhood, say $ |f(z)| \leq M $ for all $ z $ sufficiently close to~$ c $. Let the Laurent series of $ f $ be
$$
\sum_{n = -\infty}^\infty a_n(z - c)^n,
$$
so that
$$
a_n = \dfrac{1}{2\pi i} \int_\gamma \dfrac{f(\zeta)}{(\zeta - c)^{n + 1}}\,d\zeta
$$
for any anticlockwise circle~$ \gamma $ with centre~$ c $ and with sufficiently small radius~$ r $. If $ \zeta $~is in~$ \gamma $, then
$$
\left|\dfrac{f(\zeta)}{(\zeta - c)^{n + 1}}\right| = \dfrac{|f(\zeta)|}{r^{n + 1}} \leq \dfrac{M}{r^{n + 1}},
$$
and the length of~$ \gamma $ is $ 2\pi r $, so it follows from Proposition~\ref{CMT6.3} that
$$
|a_n| \leq \dfrac{1}{2\pi} \dfrac{M}{r^{n + 1}}(2\pi r) = Mr^{-n}.
$$
If $ n < 0 $, then $ r^{-n} \to 0 $ as $ r \to 0 $, so $ a_n = 0 $. Therefore the singularity is removable.

(2) If the singularity is a pole, then we obviously have $ |f(z)| \to \infty $ as $ z \to c $. Indeed, if $ c $ is a pole of order $ N $ then
we can write
$$
f(z) = \sum_{n = -N}^\infty a_n(z - c)^n = (z - c)^{-N}[a_{-N} + a_{-N + 1}(z - c) + \cdots\,]
$$
and since $ \lim_{z \to c} (a_{-N} + a_{-N + 1}(z - c) + \cdots\,) = a_{-N} $, the behaviour of $ f $ near $ c $ is determined by the factor $ (z - c)^{-N} $.

Conversely, suppose that $ |f(z)| \to \infty $ as $ z \to c $. Then there is an open disc~$ D $ with centre~$ c $ such that $ |f(z)| > 1 $ for $ z \in D \setminus\{c\} $. It
follows that $ 1/f $~is defined on $ D \setminus\{c\} $ and is bounded on that set, so $ 1/f $ has a removable singularity at~$ c $ by part~(1). Since $ 1/f(z) \to 0 $ as $ z \to c $,
and since $ 1/f(z) \neq 0 $ for $ z \neq c $, looking at the Taylor series of $ 1/f $ at~$ c $ shows that there exists $ N > 0 $ such that
$$
\dfrac{1}{f(z)} = (z - c)^N[a_N + a_{N + 1}(z - c) + \cdots\,]
$$
with $ a_N \neq 0 $. Therefore $ 1/f(z) = (z - c)^N g(z) $ with $ g $ holomorphic and non-zero near~$ c $, hence $ f(z) = (z - c)^{-N} h(z) $ with $ h $ holomorphic
and non-zero near~$ c $. Since $ N > 0 $ we conclude that $ f $~has a pole of at~$ c $.

(3) This follows from parts (1)~and~(2).
\end{proof}

\vfill\pagebreak



\subsection{The calculus of residues} \label{CMS12}

Assume $ f $ is a holomorphic function on an open set $ U \setminus \{c\} $. In the Laurent series expansion
$$
f(z) = \sum_{n = -\infty}^\infty a_n(z - c)^n,
$$
see Theorem~\ref{CMT10.1}, the coefficient~$ a_{-1} $ has the particularly simple expression
$$
a_{-1} = \dfrac{1}{2\pi i} \int_\gamma f(z)\,dz,
$$
where $ \gamma $~is a suitable circle. This coefficient has a special name, and is of special importance for calculating integrals.

\begin{definition} \label{CMD12.1}
If a holomorphic function~$ f $ has an isolated singularity at~$ c $, then the \emph{residue} of~$ f $ at~$ c $, denoted $ \res(f,c) $, is the coefficient of $ (z - c)^{-1} $
in the Laurent series of $ f(z) $ at~$ c $.
\end{definition}

\begin{theorem}[Cauchy's residue theorem] \label{CMT12.2}
Let $ U $ be an open set in~$ \C $, let $ f: U \to \C $ be holomorphic, and let $ \gamma $ be an anticlockwise simple closed path in~$ U $ such that there are finitely many
points $ c_1, \dots, c_k $ inside~$ \gamma $ not belonging to~$ U $. Then
$$
\int_\gamma f(z)\,dz = 2\pi i[\res(f,c_1) + \cdots + \res(f,c_k)].
$$
\end{theorem}

\begin{proof}[Sketch of proof]
Let $ D_1, \dots, D_k $ be disjoint open discs inside~$ \gamma $ with their centres at $ c_1, \dots, c_k $. One can show that $ \gamma $~is homotopic in~$ U $ to a path consisting
of arcs making up the boundaries $ \gamma(j) $ of these discs traced anticlockwise and of connecting paths traced twice in opposite directions. The integrals along the connecting
paths cancel out by Proposition~\ref{CMP6.2}. It therefore follows from Theorem~\ref{CMT7.3} that
$$
\int_\gamma f(z)\,dz = \sum_{j = 1}^k \int_{\gamma(j)} f(z)\,dz = \sum_{j = 1}^k 2\pi i\res(f,c_j).
$$
This yields the claim.
\end{proof}

Theorem \ref{CMT12.2} generalises Cauchy's Theorem ~\ref{CMT7.6}, which is the case with no singularities, and Cauchy's integral formulae, see Theorems~\ref{CMT8.1} and~\ref{CMT8.2}, which are cases with one singularity.

In order to apply the residue theorem, it is necessary to calculate residues. This can sometimes be done by direct computations of Laurent series. For poles there is a useful
general result as follows.

\begin{proposition} \label{CMP12.3}
If $ f $~has a pole of order at most~$ m $ at~$ c $ then
$$
\res(f,c) = \dfrac{1}{(m - 1)!} \lim_{z \to c} \dfrac{d^{m - 1}}{dz^{m - 1}}[(z - c)^m f(z)].
$$
In particular, if $ f $~has a simple pole at~$ c $ then
$$
\res(f,c) = \lim_{z \to c}[(z - c)f(z)].
$$
\end{proposition}

\begin{proof}
Let $ g(z) = (z - c)^m f(z) $, so that $ g $~is holomorphic at~$ c $, and let the Taylor series of~$ g $ at~$ c $ be $ a_0 + a_1(z - c) + \cdots\, $. The Laurent series of~$ f $
is then given by
\begin{align*}
f(z) &= \dfrac{1}{(z - c)^m}g(z) \\
&= \dfrac{1}{(z - c)^m}[a_0 + a_1(z - c) + \cdots + a_{m - 1}(z - c)^{m - 1} + \cdots\,] \\
&= a_0(z - c)^{-m} + a_1(z - c)^{-m + 1} + \cdots + a_{m - 1}(z - c)^{-1} + \cdots,
\end{align*}
so that that $ \res(f,c) = a_{m - 1} $. Therefore
$$
\res(f,c) = \dfrac{g^{(m - 1)}(c)}{(m - 1)!} = \lim_{z \to c} \dfrac{g^{(m - 1)}(z)}{(m - 1)!}
= \dfrac{1}{(m - 1)!} \lim_{z \to c} \dfrac{d^{m - 1}}{dz^{m - 1}}[(z - c)^m f(z)]
$$
as claimed.
\end{proof}

Proposition \ref{CMP12.3} can sometimes be combined with l'H\^opital's rule.

\begin{proposition}[l'H\^opital's rule] \label{CMP12.4}
If $ f $~and~$ g $ are holomorphic function on an open set $ U \subset \C $ and $ f(c) = 0 = g(c) $ for some $ c \in U $ then
$$
\lim_{z \to c} \dfrac{f(z)}{g(z)} = \lim_{z \to c} \dfrac{f'(z)}{g'(z)}.
$$
\end{proposition}

Sometimes it is possible to use partial fractions to calculate residues.

\begin{example} \label{CME12.5}
Find the residues of the following functions at their singularities:
\begin{align*}
f(z) &= \dfrac{1}{z^3 - 1}, \\
g(z) &= \dfrac{z^2 + 5z + 3}{z(z + 1)^2}, \\
h(z) &= \dfrac{\sin(z)}{z^6}, \\
k(z) &= \dfrac{1}{z - 1} + \dfrac{3z + 1}{(z - 2)^2} + \dfrac{3}{z - 4}.
\end{align*}
\end{example}

\begin{solution}
The singularities of~$ f $ are simple poles at $ 1, e^{2\pi i/3}, e^{4\pi i/3} $. For each singularity~$ c $, by Proposition~\ref{CMP12.3},
$$
\res(f,c) = \lim_{z \to c} \dfrac{z - c}{z^3 - 1}.
$$
By l'H\^opital's rule,
$$
\res(f,c) = \lim_{z \to c} \dfrac{\dfrac{d}{dz}(z - c)}{\dfrac{d}{dz}(z^3 - 1)} = \lim_{z \to c} \dfrac{1}{3z^2} = \dfrac{1}{3c^2}.
$$

The singularities of~$ g $ are a simple pole at~$ 0 $ and a double pole at~$ -1 $. By Proposition~\ref{CMP12.3},
$$
\res(g,0) = \lim_{z \to 0} zg(z) = 3
$$
and
$$
\res(g,-1) = \lim_{z \to -1} \dfrac{d}{dz} \dfrac{z^2 + 5z + 3}{z} = \lim_{z \to -1}(1 - 3z^{-2}) = -2.
$$

The only singularity of~$ h $ is at~$ 0 $. The Laurent series at~$ 0 $ is
$$
z^{-6}\left(z - \dfrac{z^3}{3!} + \dfrac{z^5}{5!} - \cdots\,\right) = z^{-5} - \tfrac{1}{6} z^{-3} + \tfrac{1}{120} z^{-1} - \cdots,
$$
so $ \res(h,0) = \tfrac{1}{120} $.

The singularities of~$ k $ are at $ 1,2 $ and~$ 4 $. The Laurent series can be found by adding the Laurent series for the individual partial fractions. The terms involving
negative powers in the Laurent series are
\begin{align*}
&(z - 1)^{-1}, \\
&\dfrac{3z + 1}{(z - 2)^2} = \dfrac{3(z - 2) + 7}{(z - 2)^2} = 7(z - 2)^{-2} + 3(z - 2)^{-1}, \\
&3(z - 4)^{-1},
\end{align*}
so $ \res(k,1) = 1, \res(k,2) = 3, \res(k,4) = 3 $.
\end{solution}

\begin{example} \label{CME12.6}
Evaluate
$$
\int_\gamma k(z)\,dz,
$$
where
$$
k(z) = \dfrac{1}{z - 1} + \dfrac{3z + 1}{(z - 2)^2} + \dfrac{3}{z - 4}
$$
and $ \gamma $~is the boundary of the square with vertices at $ \pm 3 \pm 3i $ traced anticlockwise.
\end{example}

\begin{solution}
The singularities of~$ k $ inside~$ \gamma $ are at $ 1 $~and~$ 2 $, so
$$
\int_\gamma k(z)\,dz = 2\pi i[\res(k,1) + \res(k,2)] = 2\pi i(1 + 3) = 8\pi i,
$$
using the results from Example \ref{CME12.5}.
\end{solution}

\begin{remark} \label{CMR12.7}
There is a generalisation of the residue theorem to arbitrary closed paths with essentially the same proof. If $ \gamma $~is any closed path in~$ \C $ and if $ c $~is a point
not in the image of~$ \gamma $ then there is a \emph{winding number}
$$
w(\gamma,c) = \dfrac{1}{2\pi i} \int_\gamma \dfrac{1}{z - c}\,dz,
$$
which is an integer and is the number of times that $ \gamma $~goes round~$ c $ in the  anticlockwise direction. If $ U $~is open in~$ \C $, if $ f $~is holomorphic on~$ U $,
if $ \gamma $~is a closed path in~$ U $, and if there are finitely many points $ c_1, \dots, c_k $ with non-zero winding number not belonging to~$ U $, then
$$
\int_\gamma f(z)\,dz = 2\pi i[w(\gamma,c_1)\res(f,c_1) + \cdots + w(\gamma,c_k)\res(f,c_k)].
$$
\end{remark}

\vfill\pagebreak


\subsection{Applications to integrals over real intervals} \label{CMS13}

The calculus of residues can be used to evaluate certain integrals of real functions over real intervals. One example is the famous Gaussian integral
$$
\int_{-\infty}^\infty e^{-x^2}\,dx=\sqrt{\pi},
$$
but we will not explain how to derive this formula using residues here.

We will consider integrals of two other types. Firstly, integrals of the form
$$
\int_0^{2\pi} f(\theta)\,d\theta,
$$
such that $ f $~is periodic with period $ 2\pi $ can sometimes be evaluated by the substitution $ z = e^{i\theta} $.

\begin{example} \label{CME13.1}
Compute the integral
$$
I = \int_0^{2\pi} \dfrac{1}{5 - 4\cos(\theta)}\,d\theta.
$$
\end{example}

\begin{solution}
Make the change of variable $ z = e^{i\theta} $, so that
\begin{align*}
\cos(\theta) &= \tfrac{1}{2}(e^{i\theta} + e^{-i\theta}) = \tfrac{1}{2}(z+z^{-1}), \\
dz &= ie^{i\theta} d\theta = iz\,d\theta, \\
d\theta &= -i\,\dfrac{dz}{z}.
\end{align*}
Let $ \gamma $ be the circle with centre~$ 0 $ and radius~$ 1 $ traced anticlockwise; then
$$
I = \int_\gamma\dfrac{-i}{5 - 2(z + z^{-1})} \dfrac{dz}{z} = \int_\gamma \dfrac{i}{2(z - \tfrac{1}{2})(z - 2)}\,dz.
$$
The integrand is holomorphic on~$ \gamma $ and has one singularity inside~$ \gamma $, at~$ \tfrac{1}{2} $. The residue at~$ \tfrac{1}{2} $ is
$$
\dfrac{i}{2(\tfrac{1}{2} - 2)} = -\tfrac{1}{3}i,
$$
see Proposition~\ref{CMP12.3}. Therefore
$$
I = 2\pi i(-\tfrac{1}{3}i) = \tfrac{2}{3}\pi
$$
by the residue theorem.
\end{solution}

Let $ f $ be a function such that $ z^2 f(z) $ is bounded for large values of~$ |z| $. The integral
$$
\int_{-\infty}^\infty f(x)\,dx
$$
is convergent by comparison with
$$
\int_{-\infty}^\infty \dfrac{1}{1 + x^2}\,dx = \lim_{R,S \to \infty}[\tan^{-1}x]_{-S}^R = \pi;
$$
in fact
$$
\int_{-\infty}^\infty f(x)\,dx = \lim_{R \to \infty} \int_{-R}^R f(x)\,dx.
$$
This limit can sometimes be computed by integrating round the half-discs
$$
\{z \mid |z| \leq R,\ \Im(z) \geq 0 \};
$$
the boundedness condition on $ z^2 f(z) $ ensures that $ z f(z) \to 0 $ as $ |z| \to \infty $, and from this it follows that the integrals along the semicircular parts of the
boundaries of the half-discs tend to zero.

\begin{example} \label{CME13.2}
Compute the integral
$$
I = \int_{-\infty}^\infty \dfrac{1}{(x^2 + 1)(x^2 + 4)}\,dx.
$$
\end{example}

\begin{solution}
Let
$$
f(z) = \dfrac{1}{(z^2 + 1)(z^2 + 4)} = \dfrac{1}{(z + i)(z - i)(z + 2i)(z - 2i)}.
$$
For $ R > 2 $, let
$$
I_R = \int_\gamma f(z)\,dz,
$$
where $ \gamma $~is the boundary of the upper half of the disc with centre~$ 0 $ and radius~$ R $. There are no singularities of~$ f $ on~$ \gamma $, and the singularities
inside~$ \gamma $ are at $ i $~and~$ 2i $. By Proposition~\ref{CMP12.3},
\begin{align*}
\res(f,i) &= \dfrac{1}{(i + i)(i + 2i)(i - 2i)} = -\tfrac{1}{6}i, \\
\res(f,2i) &= \dfrac{1}{(2i + i)(2i - i)(2i + 2i)} = \tfrac{1}{12}i,
\end{align*}
therefore we get
$$
I_R = 2\pi i(-\tfrac{1}{6}i + \tfrac{1}{12}i) = \tfrac{1}{6}\pi.
$$

The semicircular part of~$ \gamma $ has length $ \pi R $; hence by Proposition~\ref{CMT6.3} the absolute value along the semicircular part of~$ \gamma $ is at most
$$
\pi R \max_{|z| = R} |f(z)|.
$$
This clearly tends to zero as $ R $~tends to infinity, because $ |f(z)| $~is approximately $ 1/R^4 $ for $ |z| = R $ when $ R $~is large. The integral along the straight line
part of~$ \gamma $ is
$$
\int_{-R}^R f(x)\,dx,
$$
which tends to~$ I $ as $ R $~tends to infinity, because $ z^2 f(z) $ is bounded for large values of~$ |z| $. Therefore we get
$$
I = \tfrac{1}{6}\pi
$$
for the value of the integral.
\end{solution}

In examples of this kind, students in Methods in complex analysis must prove that the integral along the semicircle tends to zero as the radius tends to infinity; students
in Mathematics~3U may assume this without proof.

\vfill\pagebreak


\section*{PART III. CONFORMAL TRANSFORMATIONS}

\subsection{Introduction to conformal transformations} \label{CMS14}

Let $ f: U \to V $ be a bijective function between open sets in~$ \C $. We say that $ f $~is \emph{conformal} if $ f $~preserves oriented angles between (the tangents to)
intersecting curves. It turns out that this condition is equivalent to $ f $ being holomorphic with everywhere non-zero derivative.

A function from~$ \C $ to~$ \C $ is an orientation-preserving isometry if and only if it is a composite of rotations and translations. Analogously, one can show
that a function from~$ \C $ to~$ \C $ preserves oriented angles between straight lines if and only if it is a composite of rotations, translations and \emph{scalings}, where a scaling
is a function of the form
$$
z \mapsto Kz
$$
for some positive real constant~$ K $. Given arbitrary points $ c $~and~$ d $, one can deduce that the functions from~$ \C $ to~$ \C $ preserving oriented angles between straight
lines and taking~$ c $ to~$ d $ are the functions of the form
$$
z \mapsto d + Ke^{i\phi}(z - c)
$$
with $ K,\phi \in \R $, and with $ K > 0 $. In other words, they are the functions
$$
z \mapsto d + k(z - c)
$$
with $ d,k \in \C $ and with~$ k \neq 0 $.

Conformal functions are functions of approximately this form in the neighbourhood of each given point; in other words, they are the holomorphic functions with everywhere
non-zero derivatives.
%Roughly speaking, if a non-constant holomorphic function has a zero derivative at a point~$ c $, then the Taylor series at~$ c $ has the form
%$$
%a_0 + a_N(z - c)^N + a_{N + 1}(z - c)^{N + 1} + \cdots
%$$
%with $ N > 1 $ and $ a_N \neq 0 $, so the function is close to
%$$
%z \mapsto a_0 + a_N(z - c)^N
%$$
%for $ z $ close to~$ c $, and this approximation is not injective.

One can show that a holomorphic function with non-zero derivative is locally a bijection. That is, if $ f $ is holomorphic and $ f'(c) \neq 0 $ for some $ c $ then
there exists an open neighborhood $ U $ of $ c $ and an open set $ V \subset \mathbb{C} $ such that $ f: U \rightarrow V $ is a bijection.
We give the following definition.

\begin{definition} \label{CMD14.1}
Let $ U $~and~$ V $ be open subsets of~$ \C $. A \emph{conformal transformation} from~$ U $ to~$ V $ is a bijective holomorphic function from~$ U $ to~$ V $.
One says that $ U $~and~$ V $ are \emph{conformally equivalent} if there is a conformal transformation from~$ U $ to~$ V $.
\end{definition}

One can show that an injective holomorphic function has everywhere non-zero derivatives.
Since injective holomorphic functions have non-zero derivatives we have the following result.

\begin{proposition} \label{CMP14.2}
If $ f: U \to  V $ is a conformal transformation, then $ f'(c) \neq 0 $ for all $ c \in U $.
\end{proposition}

Recall that an equivalence relation on a set $ X $ is a binary relation which is reflexive, symmetric, and transitive.
If we write the relation in the form $ x \sim y $ for $ x,y \in X $ then this means

\begin{itemize}
\item $ x \sim x $ for all $ x \in X $ (Reflexivity)
\item $ x \sim y \iff y \sim x $ for all $ x,y \in X $ (Symmetry)
\item $ x \sim y, y \sim z \implies x \sim z $ for all $ x,y,z \in X $ (Transitivity).
\end{itemize}

Let us verify that conformal equivalence is indeed an equivalence relation in a natural way.

\begin{proposition} \label{CMP14.3}
Conformal equivalence is an equivalence relation on the set of all open subsets of~$ \C $.
\end{proposition}

\begin{proof}
Let $ U $ be an open subset of~$ \C $. Then the identity is a conformal transformation from~$ U $ to~$ U $, so $U $~is conformally equivalent to~$ U $. This shows
reflexivity.

To check symmetry suppose that $ U $~is conformally equivalent to~$ V $, so that there is a bijective holomorphic function $ f: U \to V $. The inverse $ g: V \to U $ is a
bijection such that
$$
f\bigl(g(y)\bigr) = y
$$
for all $ y \in V $. Differentiating this equation implicitly gives
$$
f'\bigl(g(y)\bigr)g'(y) = 1,
$$
so that $ g $~is differentiable with
$$
g'(y) = f'\bigl(g(y)\bigr)^{-1};
$$
this argument is valid because $ f'\bigl(g(y)\bigr) \neq 0 $ by Proposition~\ref{CMP14.2}. Therefore $ V $~is conformally equivalent to~$ U $.

Finally, to verify transitivity suppose that $ U $~is conformally equivalent to~$ V $ and that $ V $~is conformally equivalent to~$ W $, so that there are conformal transformations
$$
f: U \to V,\quad g: V \to W.
$$
It follows from the chain rule that $ g \circ f $ is a conformal transformation from~$ U $ to~$ W $; therefore $ V $~is conformally equivalent to~$ W $.

We conclude that conformal equivalence therefore has the three properties required for an equivalence relation.
\end{proof}

If $ f:U \to  V $ is a conformal transformation, then the holomorphic functions on~$ U $ are the functions $ g \circ f $ such that $ g $~is holomorphic on~$ V $,
hence the harmonic functions on~$ U $ are the functions $ h \circ f $ such that $ h $~is harmonic on~$ V $, see Section~\ref{CMS5}. One can therefore, for instance, analyse fluid
flows round a subset of~$ U $ in terms of flows round the possibly simpler image in~$ V $.

\begin{example} \label{CME14.4}
Suppose that
$$
w = z + \dfrac{1}{z}.
$$
Then the exterior of a circle passing through~$ -1 $ and with radius greater than~$ 2 $ in the $ z $-plane corresponds to the exterior of a set in the $ w $-plane resembling
the outline of an aeroplane wing and called a \emph{Joukowski aerofoil}. It has a cusp at the left and a vertical tangent at the right. The top is convex. The bottom is concave
at the left and convex at the right. This idea is used in the design of aeroplanes.
\end{example}

\begin{example} \label{CME14.5}
Find conformal transformations between the sets
\begin{align*}
H &= \{z \mid \Im(z) > 0\}, \\
D &= \{z \mid |z| < 1\}, \\
W_\alpha &= \{z \mid 0 < \arg(z) < \alpha\} \quad (0 < \alpha \leq 2\pi), \\
S &= \{z \mid 0 < \Im(z) < 2\pi\}.
\end{align*}
\end{example}

\begin{solution}
There is a conformal transformation $ \theta $ from~$ H $ to~$ D $ given by
$$
z \mapsto \theta(z) = \dfrac{z - i}{z + i},
$$
with inverse
$$
w \mapsto \dfrac{i(1 + w)}{1 - w}.
$$
Note that $ \theta $ is well-defined since
\begin{align*}
|\theta(z)| < 1
&\iff |z - i| < |z + i| \\
&\iff |z - i|^2 < |z + i|^2 \\
&\iff x^2 + (y - 1)^2 < x^2 + (y + 1)^2 \\
&\iff -2y < 2y \\
&\iff y > 0
\end{align*}
where $ z = x + i y $.

There is a conformal transformation from~$ H $ to~$ W_\alpha $ given by
$$
re^{i\theta} \mapsto r^{\alpha/\pi} e^{i\alpha\theta/\pi} \quad (r > 0, 0 < \theta < \pi),
$$
corresponding to a choice of $ z \mapsto z^{\alpha/\pi} $.

There is a conformal transformation from~$ S $ to~$ W_{2\pi} $ given by $ z \mapsto e^z $.
\end{solution}

\begin{example} \label{CME14.6}
Find a conformal transformation from~$ A $ to~$ B $, where
\begin{align*}
A &= \{z \mid 0 < \Re(z) < 1\}, \\
B &= \{z \mid \Im(z) > 0\}.
\end{align*}
\end{example}

\begin{solution}
Note that $ z \in B $ if and only if $ 0 < \arg(z) < \pi $. The exponential function converts bounds on imaginary parts into bounds on arguments, because $ \Im(z) $
is the argument of $ e^z $. The function $ z \mapsto iz $ converts bounds on real parts into bound on imaginary parts. Thus
\begin{align*}
0 < \Re(z) < 1
&\iff 0 < \Im(iz) < 1 \\
&\iff 0 < \Im(\pi iz) < \pi \\
&\iff 0 < \arg(e^{\pi iz}) < \pi,
\end{align*}
and it follows that the transformation $ z \mapsto e^{\pi iz} $ maps~$ A $ bijectively onto~$ B $.
\end{solution}

\begin{theorem}[Riemann mapping theorem] \label{CMT14.7}
Let $ U $ be an open set in~$ \C $. Then $ U $~is conformally equivalent to an open disc if and only if $ U \neq \C $ and $ U \neq \emptyset $ and all closed paths in~$ U $
are homotopic in~$ U $.
\end{theorem}

A space in which all closed paths are homotopic is said to be \emph{simply connected}.

\begin{proof}[Proof of necessity]
Suppose that $ U $~is conformally equivalent to an open disc.

By Liouville's theorem, see Theorem~\ref{CMT9.2}, any holomorphic function on~$ \C $ must be constant or unbounded, so its image cannot be an open disc; therefore $ U \neq \C $.

It is obvious that $ U \neq \emptyset $.

It is clear that any two closed paths in a disc are homotopic; therefore the same must be true of~$ U $.
\end{proof}

The proof of sufficiency is much more difficult, and will not be covered in this course.

\vfill
\pagebreak

%\end{document}

\subsection{M\"obius transformations} \label{CMS15}

A \emph{M\"obius transformation} is a function of the form
$$
z \mapsto \dfrac{az + b}{cz + d},
$$
where $ a,b,c,d $ are complex constants such that $ ad - bc \neq 0 $. The condition $ ad - bc \neq 0 $ ensures that this function is a bijection from its domain to its image; the
inverse is given by
$$
w \mapsto \dfrac{dw - b}{-cw + a}.
$$
If $ c = 0 $ we get a bijection from~$ \C $ to~$ \C $; if $ c \neq 0 $ we get a bijection from $ \C \setminus \{-d/c\} $ to $ \C \setminus \{a/c\} $; in any case we get a
conformal transformation. It is convenient to regard it as a bijection from $ \C \cup \{\infty\} $ to $ \C \cup \{\infty\} $ by making the definitions
$$
-d/c \mapsto \infty, \quad \infty \mapsto a/c
$$
(meaning $ \infty \mapsto \infty $ in the case $ c = 0 $). With this convention, M\"obius transformations can be composed: if
$$
\theta_1(z) = \dfrac{a_1z + b_1}{c_1z + d_1}, \quad \theta_2(z) = \dfrac{a_2z + b_2}{c_2z + d_2},
$$
then
$$
\theta_1 \theta_2(z) = \dfrac{a_3z + b_3}{c_3z + d_3},
$$
where
$$
\begin{bmatrix}
a_1 & b_1 \\
c_1 & d_1
\end{bmatrix}
\begin{bmatrix}
a_2 & b_2 \\
c_2 & d_2
\end{bmatrix}
=
\begin{bmatrix}
a_3 & b_3 \\
c_3 & d_3
\end{bmatrix}.
$$

\begin{proposition} \label{CMP15.1}
Under a M\"obius transformation, the image of a set which is a circle or a straight line is again a set which is a circle or a straight line (with the convention
that $ \infty $~belongs to every straight line).
\end{proposition}

\begin{proof}
The original set consist of the points~$ z $ satisfying a condition of the form
$$
\dfrac{|z - p|}{|z - q|} = k,
$$
where $ p $~and~$ q $ are distinct complex numbers and where $ k $~is a positive real number. More precisely, for a straight line one needs $ k = 1 $,
and for a circle one needs $ k \neq 1 $. Making the substitution $ z = (dw - b)/(-cw + a) $ gives a condition of the same form for~$ w $. Indeed,
we calculate
\begin{align*}
k = \dfrac{|z - p|}{|z - q|} &= \dfrac{\big|\frac{dw - b}{-cw + a} - p\big|}{\big|\frac{dw - b}{-cw + a} - q\big|} \\
&= \dfrac{|dw - b + p(cw - a)|}{|dw - b + q(cw - a)|} \\
&= \dfrac{|(d + pc)w - pa - b|}{|(d + qc)w - qa - b|},
\end{align*}
and multiplying both sides with $ (d + qc)/(c + pc) $ gives a relation of the required form for $ w $.
%Need to be careful here, since we may be dividing by zero. If the denominator is zero then we get a circle anyway though.
\end{proof}

\begin{proposition} \label{CMP15.2}
Let $ D $ be the open disc with centre~$ 0 $ and radius~$ 1 $, and let $ f: D \to D $ be a conformal transformation such that $ f(0) = 0 $. Then $ f(z) = \kappa z $ for
some constant~$ \kappa $ with $ |\kappa| = 1 $.
\end{proposition}

\begin{proof}[Sketch of proof]
Consider the function $ f(z)/z $, which is holomorphic on $ D \setminus \{0\} $. We have $ f(z)/z \to f'(0) $ as $ z \to 0 $, so the singularity at~$ 0 $ is
removable (see Theorem~\ref{CMT11.5}). Thus there is a holomorphic function~$ g $ on~$ D $ such that
$$
g(z) = \dfrac{f(z)}{z}
$$
for $ z \neq 0 $. We want to show that $ g(z) $ is a constant with absolute value~$ 1 $.
Let us first prove that $ |g(z)| = 1 $ for all $ z \in D $.

For $ 0 < r < 1 $ one can show that the maximum value of $ |g(z)| $ with $ |z| \leq r $ is attained at some point~$ z_0 $ with $ |z_0| = r $, see Question 5 on Exercise Sheet 5.
%Excercise 31.
For $ |z| \leq r $ it follows that
$$
\dfrac{|f(z)|}{|z|} = |g(z)| \leq |g(z_0)| = \dfrac{|f(z_0)|}{|z_0|} \leq \dfrac{1}{r},
$$
using $ f(z_0) \in D $ in the last inequality.
Letting $ r \to 1 $ shows that $ |f(z)| \leq |z| $ for $ z \in D $. Applying the same argument to~$ f^{-1} $ shows that $ |z| \leq |f(z)| $ for all $ z \in D $, hence $ |g(z)| = 1 $
for all $ z \in D $ as claimed.

We can now show that $ g(z) = \kappa $ for some $ \kappa \in \C $ with $ |\kappa| = 1 $. Decompose $ g = u + iv $ into its real and imaginary parts. Then from
$ |u + iv| = 1 $ we get $ u^2 + v^2 = 1 $ which implies
$$
u \frac{\partial u}{\partial x} + v \frac{\partial v}{\partial x} = 0, \qquad u \frac{\partial u}{\partial y} + v \frac{\partial v}{\partial y} = 0.
$$
Inserting the Cauchy--Riemann equations, see Theorem~\ref{CMT4.1}, into the second equality gives
$$
- u \frac{\partial v}{\partial x} + v \frac{\partial u}{\partial x} = 0.
$$
%give $ \partial_x u = \partial_y v, \partial_y u = - \partial_x v $,
Multiplying this and the first equality by $ u $ and $ v $, respectively, and adding up the results shows
$$
(u^2 + v^2)\frac{\partial u}{\partial x} = \frac{\partial u}{\partial x} = 0, \qquad (u^2 + v^2)\frac{\partial v}{\partial x} = \frac{\partial v}{\partial x} = 0.
$$
It follows that
$$
\frac{\partial u}{\partial x} = 0 =  \frac{\partial v}{\partial x},
$$
and similarly one shows that the partial derivatives of $ u, v $ with respect to $ y $ vanish. This means that $ u $ and $ v $ are constant, which finishes the proof.
\end{proof}

\begin{theorem} \label{CMT15.3}
Let $ D $ be the open disc with centre~$ 0 $ and radius~$ 1 $, and let $ f $ be a conformal transformation from~$ D $ to~$ D $. Then
$$
f(z) = \kappa\dfrac{z - a}{\bar{a} z - 1}
$$
for some constants $ \kappa $~and~$ a $ with $ |\kappa| = 1 $ and $ |a| < 1 $.
\end{theorem}

\begin{proof}[Sketch proof]
Let $ a = f^{-1}(0) $, and let
$$
h(z) = \dfrac{z - a}{\bar{a} z - 1}.
$$
Then $ h $~is a conformal transformation from~$ D $ to~$ D $ such that $ h(a) = 0 $. It follows that $ f \circ h^{-1} $ is a conformal transformation from~$ D $ to~$ D $
sending~$ 0 $ to~$ 0 $; therefore
$$
f\bigl(h^{-1}(w)\bigr) = \kappa w
$$
for some constant~$ \kappa $ with $ |\kappa| = 1 $. Putting $ w = h(z) $ gives the result.
\end{proof}

\begin{example} \label{CME15.4}
Let
$$
D = \{z \in \C \mid |z| < 1\}.
$$
Find the image of~$ D $ under the transformation
$$
z \mapsto \dfrac{1 + z}{1 - 2z}.
$$
\end{example}

\begin{solution}
We can use the idea of the proof of Proposition~\ref{CMP15.1}. Let
$$
w = \dfrac{1 + z}{1 - 2z};
$$
then
$$
z = \dfrac{w - 1}{2w + 1}.
$$
Let $ w = u + iv $ with $ u,v \in \R $; then
\begin{align*}
|z| < 1 &\iff |w - 1| < |2w + 1| \\
&\iff |u + iv - 1| < |2u + 2iv + 1| \\
&\iff (u - 1)^2 + v^2 < (2u + 1)^2 + (2v)^2 \\
&\iff u^2 + v^2 - 2u + 1 < 4u^2 + 4v^2 + 4u + 1 \\
&\iff 3u^2 + 3v^2 + 6u > 0 \\
&\iff u^2 + v^2 + 2u > 0 \\
&\iff (u + 1)^2 + v^2 > 1 \\
&\iff |w + 1| > 1.
\end{align*}
The image of~$ D $ therefore consists of the points outside the circle with centre~$ -1 $ and radius~$ 1 $.

To check this, note that
$$
-1 \mapsto 0, 1\mapsto -2, i \mapsto -1 + \tfrac{1}{5}(4 + 3i),
$$
so that the boundary circle of~$ D $ is sent to the circle through $ 0,-2 $ and $ -\tfrac{1}{5} + \tfrac{3}{5}i $. This is the circle with centre~$ -1 $ and radius~$ 1 $. It follows that
the image of~$ D $ is either the region inside the image circle or the region outside the image circle. Since $ 0 \mapsto 1 $, the image is in fact the outside.
\end{solution}

\begin{example} \label{CME15.5}
Let $ f $ be a M\"obius transformation such that
$$
f(\R \cup \{\infty\}) = \R \cup \{\infty\}.
$$
Show that
$$
f(z) = \dfrac{az + b}{cz + d}
$$
for some real constants $ a,b,c,d $.
\end{example}

\begin{solution}
We know that
$$
f(z) = \dfrac{Az + B}{Cz + D}
$$
for some complex constants $ A,B,C,D $ such that $ AD - BC \neq 0 $.

Suppose that $ C = 0 $. Then $ A \neq 0 $ and $ D \neq 0 $, so $ f(z) $ is a linear polynomial, say
$$
f(z) = az + b = \dfrac{az + b}{0z + 1}
$$
with $ a \neq 0 $. We have $ f(0) = b $ and $ f(1) = a + b $. Since $ f(0) $ and $ f(1) $ are real, the same is true for $ a $~and~$ b $.

Suppose that $ C \neq 0 $. Then we can write $ f(z) $ in the form
$$
f(z) = \dfrac{az + b}{z + d}
$$
with $ ad - b \neq 0 $. We have $ f^{-1}(\infty) = -d $; therefore $ d \in \R $. We have $ f(0) = b/d $, so that $ b = d f(0) $; therefore $ b \in \R $. Finally, we
have $ f(1) = (a + b)/(1 + d) $, so that
$$
a = (1 + d)f(1) - b;
$$
therefore $ a \in \R $.
\end{solution}


\end{document}

%For use of axessibility:

\tagpdfsetup{tabsorder=structure,uncompress,activate-all,interwordspace=true}
\tagpdfifpdftexT
{
\pdfcatalog{/Lang (en-US)}
\usepackage[T1]{fontenc}
\input glyphtounicode
\pdfgentounicode=1
}
\tagpdfifluatexT
{
\pdfextension catalog{/Lang (en-US)}
\RequirePackage{fontspec}
\RequirePackage{luacode}
\newfontface\zerowidthfont{freeserif}
\directlua{
require(axessibility.lua)
}
}
